Bacteria detect neutrophils via a system that responds to hypochlorous acid and flow

Version 1

Abstract

Abstract

Neutrophils respond to the presence of bacteria by producing oxidative molecules that are lethal to bacteria, including hypochlorous acid (HOCl). However, the extent to which bacteria detect activated neutrophils or the HOCl that neutrophils produce, has not been understood. Here we report that the opportunistic bacterial pathogen Pseudomonas aeruginosa upregulates expression of its fro operon in response to stimulated neutrophils. This operon was previously shown to be activated by shear rate of fluid flow in the environment. We show that fro is specifically upregulated by HOCl, while other oxidative factors that neutrophils produce including H2O2, do not upregulate fro. The fro-dependent response to HOCl upregulates the expression of multiple methionine sulfoxide reductases, which relieve oxidative stress that would otherwise inhibit growth. Our findings suggest a model in which the detection of shear rate or HOCl activates the fro operon, which serves as an early and sensitive host-detection system for P. aeruginosa that improves its own survival against neutrophil-mediated host defenses. In support of this model, we found that the fro operon is activated in an infection model where flow and neutrophils are present. This response could promote the bacterium’s pathogenicity, colonization of tissue, and persistence in infections.

Introduction

P. aeruginosa is a Gram-negative bacterium that is a major cause of hospital-acquired infections and is a significant antibiotic resistance threat [1]. To survive in a broad range of conditions, this bacterium has evolved signaling pathways that relay information about changes in nutrients, chemical gradients, and mechanical forces such as those generated by fluid flow, and activates responses that promote its survival [2,3]. Understanding how extracellular-sensing pathways detect and respond to host cues is critical for improving treating bacterial infections.

Mammalian host immune systems respond to the presence of bacterial pathogens by recruiting immune cells that migrate rapidly through circulating blood to sites of infection [4]. Neutrophils are the first and most abundant immune cells recruited to inflamed or damaged tissues [5] and are stimulated into an activated state by bacterial components including the membrane constituent lipopolysaccharide. Activated neutrophils produce multiple products that are toxic to bacteria including reactive oxygen species (ROS) and reactive chlorine species (RCS) [6–9]. While the toxic effects of ROS on bacteria are understood, far less is known about how bacteria are affected by RCS [10,11]. Hypochlorous acid (HOCl) is a major neutrophil-produced RCS that is formed through the activity of myeloperoxidase with H2O2 and chlorine ions [6,7,12] and is strongly bactericidal [13–15]. P. aeruginosa is a potent activator of neutrophil respiratory bursts [16] but whether and how the bacteria detect and respond to these respiratory bursts, is not understood.

P. aeruginosa detects shear rate generated by the flow of fluids via the fro operon [3,17], which is upregulated in the bacterium during infection in humans [3,18] (Fig. 1A). In addition, fro is necessary for successful colonization of the gastrointestinal tract and lung infections [19,20], although the reasons for its necessity have been unclear. Because flow is ubiquitous in the pulmonary and circulatory systems, in principle the fro operon could be upregulated in flow-associated host environments. Given its important role in infection, it is possible that Fro could also respond to other immune-associated cues, such as the presence of neutrophils. The fro operon consists of froABCD (PA14_21570, PA14_21580, PA14_21590, PA14_21600), with FroC and FroD predicted to localize to the inner membrane, and FroB predicted to localize in the cytoplasm [21]. Expression of the operon is regulated by the sigma and anti-sigma factors froR (PA14_21550) and froI (PA14_21560), respectively [22]. Overexpression of froR upregulates fro, whereas overexpression of froI downregulates fro [3]. Flow activates FroR-dependent transcription of froABCD through an unknown mechanism [3].

Activation of fro promoter transcription by flow and during infection of human tissue.

(A) RNA-seq heatmap of P. aeruginosa genes that are upregulated by at least 3-fold by flow and during infection in human wounds. Datasets are from [3,18]. (B) Schematic of a P. aeruginosa fluorescent reporter strain that contains a transcriptional fusion of yfp to the fro promoter and expresses mCherry under a constitutive promoter. (C) Representative phase contrast, fluorescence, and merged images of the P. aeruginosa reporter strain co-incubated with neutrophils, macrophages, or host cell-free medium. Dashed lines indicate boundaries of the host cells and white arrows indicate P. aeruginosa with high YFP/mCherry ratios. Images are representative of three independent experiments. Scale bars represent 5 μm.

We show that the fro operon responds directly to HOCl, which is produced by neutrophils during immune response. Using a fluorescent reporter system in live cells and through quantitative measurements of transcription, we show that fro expression is upregulated in a mouse model of corneal infection and by activated neutrophils. Through transcriptional profiling, we determine that fro upregulation mitigates the toxic effects of HOCl by activating methionine sulfoxide reductases that relieve oxidative stress in bacteria. These data suggest a novel mechanism by which bacterial defense against host innate immunity is primed by a single system to detect host defense cues through both flow and the presence of the highly potent oxidative factor HOCl.

Results

Host cells stimulate fro expression in P. aeruginosa under non-flow conditions

We investigated whether the fro operon is activated directly by immune cells. To quantify fro expression, we used a P. aeruginosa strain that co-expresses the yellow fluorescent protein (YFP) under the transcriptional control of the fro promoter, and mCherry under the control of a constitutive promoter (Fig. 1B and [3]). While fro expression increases under flow (Fig. S1A in the Supplementary Information (SI) and [3]), experiments in this study were performed in the absence of flow in order to deconvolve the effects of flow from potential activation of fro by immune cells. Neutrophils and macrophages are among the first immune cells that are recruited to sites of infection. We co-incubated P. aeruginosa with neutrophils or macrophages for 30 minutes to determine whether they would elicit a response in the fro operon. Relative to P. aeruginosa that were incubated in host cell-free media, P. aeruginosa in close proximity to human neutrophils or that were engulfed in mouse macrophages appeared to increase fro expression (Fig. 1C), suggesting that close proximity to or direct interaction with these host cells stimulate fro expression in P. aeruginosa.

Activated neutrophils induce fro expression

We focused on the potential activation of fro expression by neutrophils because P. aeruginosa triggers respiratory bursts in neutrophils [16], which generates ROS and RCS. We hypothesized that fro activation could be a response to respiratory burst products. P. aeruginosa were incubated in conditioned medium in which respiratory bursts were induced in human neutrophils using phorbol myristate acetate (PMA) [23,24] (Fig. 2A). The expression of fro was quantified by averaging the ratios of YFP to mCherry fluorescence for at least one hundred individual P. aeruginosa. In support of our hypothesis, conditioned medium from PMA-stimulated neutrophils increased expression of fro expression by 30-fold compared to conditioned medium from unstimulated neutrophils (Fig. 2B). The increased fro expression was not due to PMA alone, as PMA-treated medium without neutrophils had no effect on fro expression (Fig. 2B). Conditioned medium from unstimulated neutrophils had no significant effect on fro expression compared to medium alone (Fig. 2B), suggesting that the activation of fro is due to the products of respiratory bursts.

Conditioned medium from stimulated neutrophils induces fro expression.

(A) Representative phase-contrast and fluorescence images and (B) fro expression as determined by ratios of YFP to mCherry fluorescence, of P. aeruginosa that were cultured in medium only, medium with PMA, conditioned medium from PMA-stimulated neutrophils, or conditioned medium from unstimulated neutrophils. Scale bars represent 5 μm. Data points indicate an average from at least one hundred individual P. aeruginosa. Gray columns indicate the mean of at least three independent experiments and error bars represent the standard error of the mean (SEM). P-values were obtained using two-tailed t-tests with unequal variances and values of p>0.05 are denoted as ns (nonsignificant). (C) Schematic depicting that PMA-stimulated neutrophils produce respiratory burst products that could activate fro expression.

HOCl but not H2O2 or HNO3 induces fro expression

Activated neutrophils generate ROS and RCS through respiratory bursts (Fig. 2C), which have an inhibitory effect on bacterial growth [6,7,10,13,14]. To explore the hypothesis that these reactive species induce fro expression, we measured the effects of NaOCl, H2O2, and HNO3 on fro expression at concentrations just below the minimum inhibitory concentrations (MICs) (Fig. 3A-B). NaOCl increased fro expression by up 74-fold at 1 µM (Fig. 3B). At 2 µM, NaOCl caused only a 3-fold increase in fro expression but these bacteria were noticeably smaller in size (Fig. S1B in the SI), raising the possibility that the treatment could have interfered with transcription. The activation of fro transcription by NaOCl was further assessed using reverse transcription quantitative PCR (RT-qPCR). In support of the fro-yfp reporter findings, a 200-fold increase in fro transcription was observed using 1 µM NaOCl (Fig. 3C).

NaOCl but not H2O2 or HNO3 induces fro expression.

(A) Growth profiles as measured by optical density (OD600) of P. aeruginosa treated with NaOCl, H2O2, HNO3, or no treatment. Data points indicate a mean of at least two independent experiments and error bars indicate SEM. (B) fro expression as determined by ratios of YFP to mCherry fluorescence in P. aeruginosa treated with NaOCl, H2O2, or HNO3, or no treatment (UTR). Horizontal bars indicate the mean of at least three independent experiments and error bars indicate SEM. The dashed line indicates the average YFP/mCherry ratio from the untreated condition. (C) Abundance of froA transcripts in P. aeruginosa following treatment with 1 µM NaOCl, 4 µM H2O2, or no treatment (UTR), as measured by RT-qPCR. Measurements were normalized to 5S ribosomal RNA and the logarithm (base 2) of the fold-change in transcription was computed relative to the untreated condition. Horizontal bars indicate the average of four independent experiments and error bars indicate SEM. (D) fro expression, measured by YFP to mCherry ratio, in P. aeruginosa after treatment with 1 µM NaOCl, 1 µM NaCl, 200 µM HCl, 6 mM NaOH, or no treatment (UTR). Error bars indicate SEM. In panels B and D, data points indicate the average from at least one hundred individual P. aeruginosa. P-values were obtained using two-tailed t-tests with unequal variances, with values of p>0.05 denoted as ns.

The increased fro expression was not due to sodium or chloride ions alone, as neither NaCl nor HCl affected fro expression in the absence of NaOCl (Fig. 3D). The fro response to NaOCl was not due to basic pH, as treatment with up to 6 mM of NaOH had no significant impact on fro activation (Fig. 3D). As fro activation is not due to sodium or chloride ions, nor changes in pH, we attributed the effects to hypochlorite (OCl-), which is in equilibrium with the hypochlorous acid species (HOCl) at the pH of 7.1 used in our experiments.

Changes in fro expression were assessed using H2O2, a strong oxidizer. Small increases in fro expression were observed using the subinhibitory concentrations of 4 and 8 µM but these changes were not statistically significant compared to untreated cultures (Fig. 3A-B), and no significant changes in fro transcript abundance were observed using qRT-PCR (Fig. 3C). The fro response was also assessed using subinhibitory concentrations of HNO3, for which no change in expression was detected up to near-MIC concentrations (Fig. 3A-B). Together, these results show that Fro expression is activated by HOCl and suggest that HOCl produced by stimulated neutrophils could activate fro expression.

Corneal infection activates fro expression

P. aeruginosa is a major cause of corneal ulcers and keratitis, which can cause visual impairment and blindness. Neutrophils are recruited within hours to infected corneal epithelia with neutrophil cell density peaking at 18 hours [25]. We tested the hypothesis that fro expression is upregulated during infection using a previously-established murine model of corneal infection in which P. aeruginosa strain PAO1F is inoculated at the site of a corneal abrasion [26]. Mice corneas were harvested 2 or 20 hours post-infection and fro transcripts were quantified using RT-qPCR. Transcription was normalized for number of P. aeruginosa using 5S ribosomal RNA. We confirmed that this strain of P. aeruginosa upregulates its fro transcription in response to NaOCl but not H2O2 (Fig. S1B in the SI), consistent with our previous results. In support of the hypothesis, fro expression increased significantly by 20 hours post-infection (Fig. 4A). Expression levels showed no significant change within the first two hours following infection, which we attribute to the low level of immune cell recruitment during this period.

The expression of fro is activated during corneal infection and inhibited by methionine and antioxidants.

(A) Abundance of froA transcripts in P. aeruginosa 0, 2, or 20 hours after being inoculated on a mouse corneal abrasion, as measured by RT-qPCR. Measurements were normalized to 5S ribosomal RNA and the logarithm (base 2) of the fold-change in transcription was computed relative to the initial inoculum. Data points represent individual experiments. Horizontal bars indicate the average of at least six independent experiments and error bars represent SEM. (B) Heatmap showing P. aeruginosa gene transcripts that were upregulated by at least 4-fold in wild type compared to the ΔfroR mutant after treatment with 1 µM NaOCl and that had p-values less than 0.05 (n≥3), sorted by genomic locus. Raw values are in Table S1 in the SI. Log2(FC) indicates the log2 of the fold-change. (C) fro expression, determined by ratios of YFP to mCherry, in P. aeruginosa with no treatment (UTR), or treated with 1 µM NaOCl, or 1 µM NaOCl with methionine (Met), cysteine (Cys), or β-mercaptoethanol (βME). (D) fro expression, determined by ratios of YFP to mCherry, in P. aeruginosa after incubation with conditioned medium only (replotted from Fig. 2B), conditioned medium from stimulated neutrophils, or in conditioned medium from stimulated neutrophils with 100 µM methionine. Data points indicate an average from at least one hundred individual P. aeruginosa. Gray columns indicate the mean of at least three independent experiments and error bars represent SEM. P-values were obtained using two-tailed t-tests with unequal variances and values of p>0.05 are denoted as ns.

Methionine sulfoxide reductase upregulation requires FroR

To understand how the HOCl response is regulated by fro, we performed transcriptional profiling of wild-type (WT) P. aeruginosa and a strain containing a deletion of FroR, which is the sigma factor that is required for fro activation [3]. The genes that were most upregulated by NaOCl in WT compared to the ΔfroR mutant were in the fro operon, the phosphate-specific and pyrophosphate-specific outer membrane porin genes oprO and oprP, and the methionine sulfoxide reductase genes msrB, msrQ and msrP (Fig. 4B). The methionine sulfoxide reductases (MSRs) have an essential role in bacterial survival and defense against RCS stress. HOCl oxidizes methionine residues 100-fold more rapidly than other cellular components, forming methionine sulfoxide, which impairs protein function [27–31]. MSRs restore proper protein function by reducing methionine sulfoxide back to methionine [11,32]. Our data suggest that froR is required for the upregulation of MSRs in response to HOCl stress. It is possible that oxidized methionine increases fro expression, which in-turn increases MSR expression.

We reasoned that if oxidized methionine upregulates fro, an excess of methionine should inhibit fro activation. In support of our interpretation, methionine co-treatment with NaOCl suppressed fro expression (Fig. 4C). We hypothesized that if fro expression is activated by HOCl oxidation, antioxidants such as cysteine and β-mercaptoethanol (βME) should suppress fro activation by functioning as alternative oxidation targets. Indeed, co-treatment of NaOCl with either cysteine or βME significantly suppressed NaOCl-activated fro expression (Fig. 4C and Fig. S2 in the SI).

We considered how fro could be upregulated by stimulated neutrophils in the context of these findings. RCS produced by neutrophils could oxidize methionine in P. aeruginosa, resulting in fro upregulation. Under this interpretation, supplying excess methionine should inhibit activation. Indeed, supplementing conditioned media from PMA-stimulated neutrophils with methionine completely suppressed fro expression (Fig. 4D). Together, these data suggest a model in which fro responds to HOCl-induced oxidation of methionine by upregulating MSRs.

FroR protects P. aeruginosa against HOCl

The upregulation of MSRs is expected to improve bacterial growth against HOCl stress. We measured the growth profiles of WT and ΔfroR strains of P. aeruginosa. Treatment with 16 µM NaOCl significantly inhibited the growth of the wild-type strain (Fig. 5A) but resulted in even greater growth inhibition in the ΔfroR strain. This observation suggests that the fro operon improves P. aeruginosa tolerance to HOCl stress.

FroR improves P. aeruginosa growth against HOCl stress and regulates transcription of genes involved in antioxidant defense.

(A) Growth profiles of wild-type P. aeruginosa (WT) and ΔfroR mutants treated with 16 µM NaOCl or untreated. Data points represent an average of at least three independent experiments and error bars indicate SEM. (B-C) Statistical significance as a function of fold-change in transcript abundance in (B) WT and (C) ΔfroR P. aeruginosa after treatment with 1 µM NaOCl (n≥3). Genes in red (upregulated) and blue (downregulated) were altered by at least four-fold (raw values in Table S2-S3 in the SI). The unlabeled genes have not been previously annotated. P-values were determined using the Wald test. Greater values along the vertical axis indicate greater statistical significance and the dashed horizontal line indicates a p-value of 0.05. (D) Schematic that depicts a model in which reactive chlorine species (RCS) reacting with methionine (Met) produces oxidized methionine, which upregulates fro and in-turn increases the production of methionine sulfoxide reductases.

We considered other genes which may improve WT P. aeruginosa tolerance to HOCl. HOCl increased the expression of iron uptake genes (pvdA and pvdQ) and downregulated the expression of oxidative phosphorylation genes (cyoCDE) (Fig. 5B and Table S2 in the SI). These changes suggest that HOCl causes an iron deficiency and decreases oxidative phosphorylation. In order to compensate for decreased oxidative phosphorylation, P. aeruginosa may utilize amino acid catabolism as an alternative energy source, as evidenced by a concomitant upregulation of branched chain amino acid catabolism genes (hpd, bkdA1, bkdA2, and bkdB) by HOCl (Fig. 5B and Table S2 in the SI).

While the ΔfroR strain was sensitive to HOCl, the strain was tolerant (Fig. 5A), suggesting that additional mechanisms relieve HOCl stress. In the ΔfroR mutant, HOCl upregulated genes encoding multidrug efflux pumps (mexEF-oprN and mexXY) (Fig. 5C and Table S3 in the SI). While these pumps can relieve oxidative stress [33,34], they were not activated in the WT strain. This observation suggests that the response to HOCl may be hierarchical, in which Fro-regulated and Mex pathways could be first-line and second-line responses to HOCl stress, respectively. Supporting this interpretation, it has been reported that treatment of the WT strain using significantly higher (millimolar) HOCl concentrations upregulates mexEF-oprN [35] and the mexXY regulator mexT [36].

Discussion

Neutrophils are activated by the presence of bacteria that infect host tissue. In their activated state, neutrophils generate respiratory bursts that produce an abundance of HOCl. Whether and how bacteria respond to activated neutrophils has not been clear. We have demonstrated that P. aeruginosa responds to activated neutrophils by upregulating the transcriptional expression of the fro operon, which relieves HOCl stress. Our data suggest that the upregulation of fro could function as a bacterial defense mechanism against neutrophil attack, thus improving P. aeruginosa survival and colonization of human tissue during infection.

One of the striking features of the fro system is that it is finely tuned to activate at sub-lethal concentrations (1-2 µM) of NaOCl. Activated neutrophils produce HOCl concentrations as high as 50 µM [24]. The activation of fro at relatively low concentrations suggests that the system functions as an early and sensitive detector of HOCl. In hosts, fro may be upregulated in environments where HOCl is present in low concentrations, such as during early stages of neutrophil recruitment in the cornea or when P. aeruginosa is spatially distant from activated neutrophils. The early activation of fro could benefit P. aeruginosa by triggering a protective mechanism before HOCl reaches lethal concentrations. The ability of the Fro system to respond to the presence of fluid flow provides an additional layer of early activation, such as in the case of host circulatory systems. Shear rates that activate fro [3] could prime P. aeruginosa defenses against circulating neutrophils. Consistent with this interpretation, shear rate sensitizes Fro to H2O2 [17], which is also produced by neutrophils in response to the presence of bacteria [6].

Based on the observations that multi-drug efflux pumps only activate in the absence of FroR and at high HOCl concentrations [35], and that multiple systems repair RCS damage [1–3], the HOCl response is likely to follow a hierarchy of activation. Our data suggest a model in which the fro system is the first-line response that suppresses the toxic effects of HOCl towards methionine (Fig. 5D). FroR activates two distinct classes of reductases that repair RCS-oxidized methionine: MsrB reduces methionine sulfoxide with electrons provided by the thioredoxin system [37,38], and MsrPQ uses electrons from the respiratory chain [39]. The activation of both reductase classes enables P. aeruginosa to limit RCS stress in both the cytoplasmic and periplasmic spaces by utilizing multiple electron donor sources.

Bacteria that are exposed to low concentrations of HOCl have higher rates of horizontal gene transfer and can acquire antibiotic resistance genes more rapidly [40,41]. The fro system could thus facilitate antibiotic resistance through increasing P. aeruginosa tolerance to HOCl. Our findings thus suggest that the fro pathway could be an important defense mechanism to target in the development of novel antibiotic therapeutics. Inhibition of this pathway could increase P. aeruginosa susceptibility to neutrophil-mediated killing while minimizing antibiotic resistance.

Materials and methods

Strains, Growth Conditions, and Reagents

Experiments were performed using the P. aeruginosa strain AL143, which is strain PA14 strain that contains the yfp gene integrated between froA and froB and a constitutively-expressed mCherry gene [3], ΔfroR and ΔfroI mutant strains that were previously constructed [3], and P. aeruginosa strain PAO1F[42,43]. Strains were streaked onto LB-Miller (BD Biosciences, Franklin Lakes, NJ) petri dishes containing 2% Bacto agar (BD Biosciences, Franklin Lakes, NJ), incubated overnight at 37 °C, and single colonies were inoculated into 2mL of modified MinA minimal medium (60.3mM K2HPO4, 33.0mM KH2PO4, 7.6mM (NH4)2SO4, 1.0mM MgSO4,) [44] containing 0.2% (w/v) sodium citrate as the carbon source. Strains were cultured in MinA minimal medium for 18 hours in a roller drum spinning at 200 rpm at 37 °C, diluted 1:1000 into fresh MinA minimal medium supplemented with oxidative or antioxidative agents, or were not supplemented and imaged using fluorescence microscopy. P. aeruginosa cultures were incubated with NaOCl, H2O2, HNO3 for 3 hours unless otherwise stated at 37 °C before imaging.

Neutrophils were isolated from blood samples from healthy donors between 18 to 65 years old that were collected at the Institute for Clinical and Translational Science at the University of California, Irvine, through approved protocols with the UCI Institutional Review Board HS#2001-2058, and informed consent was obtained from all donors. Neutrophils were prepared as described in [45]. Briefly, 3% dextran (from Leuconostoc spp, Mr 450,000-650,000, Sigma-Aldrich, St. Louis, MO) in PBS (Gibco, ThermoFisher, Waltham, MA) was used to separate red blood cells (RBCs) from whole blood. Neutrophils were purified from the remaining cells by overlaying on a Ficoll density gradient (GE Healthcare, Chicago, IL) following centrifugation in Ficoll-Paque Centrifugation Media (GE Healthcare, Chicago, IL) for 25 minutes at 500 × g. The remaining RBCs were lysed using RBC Lysis Buffer (Fisher Scientific, Hampton, NH), and neutrophils were resuspended in RPMI 1640 medium (ATCC, Manassas, VA) and purity was assessed by flow cytometry using the ACEA NovoCyte Flow Cytometer and fluorescent antibodies APC-CD11b, FITC-CD16 and PE-CD66b (eBioscience, San Diego, CA). Neutrophils were immediately used for neutrophil stimulation or co-incubation experiments with P. aeruginosa. This procedure routinely yielded greater than 95% CD11B-positive neutrophil populations.

J774.1 mouse macrophages (ATCC) were cultured in 10 mL DMEM (Gibco, ThermoFisher) in T75 flasks (VWR, Radnor, PA) at 37 °C with 5% CO2. Macrophages were passaged every three days through scraping and were passaged every 5 days through trypsinization.

For P. aeruginosa size analysis experiments, cultures were incubated at 37 °C in 250 mL flasks for 3 hours with shaking at 225 rpm following the 1:1000 dilution. For methionine and antioxidant treatments, following the 1:1000 dilution, cultures were supplemented with NaOCl and the addition of methionine, β-mercaptoethanol (βME) or cysteine. 50 mL of each bacterial culture was incubated at 37 °C in 250 mL flasks for 3 hours with shaking at 225 rpm, imaged, and analyzed as described in the fluorescence microscopy section.

Fluorescence and Phase Contrast Microscopy

Bacteria were immobilized on 1% agarose pads containing minimal medium and imaged immediately. Cultures containing densities below 10 P. aeruginosa per frame were concentrated using a syringe filter with 0.2 or 0.8 µm pore sizes (Millipore, Burlington, MA). Images were acquired using a Nikon Eclipse Ti-E microscope (Nikon, Melville, NY) containing a Nikon 100X Plan Apo (1.45 N.A.) objective, a Nikon Ph3 phase contrast condenser annulus, a Sola light engine (Lumencor, Beaverton, OR), an LED-DA/FI/TX filter set (Semrock, Rochester, NY) for visualizing the mCherry fluorescence spectrum containing a 409/493/596 dichroic and 575/25 nm filters for excitation and 641/75 nm filters for emission, an LED-CFP/YFP/MCHERRY filter set (Semrock) for visualizing YFP fluorescence containing a 459/526/596 dichroic and 509/22 nm filters for excitation and 544/24 nm filters for emission, and a Hamamatsu Orca Flash 4.0 V2 camera (Hamamatsu, Bridgewater, NJ).

Images were acquired using Nikon NIS-Elements and analyzed using custom built software written previously [46] in Matlab (Mathworks, Natick, MA). See the “Data, code, and materials availability” section below to download code. Briefly, P. aeruginosa masks were determined from phase contrast images using an edge-detection algorithm. For size analysis, the total pixel area of each individual P. aeruginosa was determined by computing the mask area and converting from pixels to μm2 by multiplying the mask area by a factor of 0.004225 μm2/pixel to account for the microscope camera pixel size and objective magnification. Fro expression was quantified by averaging computed ratios of YFP to mCherry fluorescence for at least one hundred individual P. aeruginosa. In each experiment, a minimum of 100 P. aeruginosa were imaged and used for analysis in which ratios were computed from the masked areas. To image P. aeruginosa under flow, microfluidic devices were fabricated using standard soft lithography techniques as established previously [47] with channel dimensions of 100 μm x 50 μm (width x height) and a flow rate of approximately 5 μL/min.

Co-incubation of Mammalian Cells with P. aeruginosa

P. aeruginosa were cultured for 18 hours in minimal medium, centrifuged at 4600 x g for 2 minutes and resuspended in DPBS (Gibco, ThermoFisher) three times, and diluted to a final OD600 of 0.2. Macrophages that were harvested from growth flasks or freshly prepared neutrophils were washed three times by centrifuging at 300 x g for 5 minutes and resuspending in DPBS, and 1 mL aliquots containing 4×105 cells/mL were transferred into flat-bottom dishes. P. aeruginosa culture was added to the cells at a multiplicity of infection of 10 and incubated at 37 °C with CO2 for 30 minutes. Dishes were aspirated until approximately 20 µL of medium remained and pads consisting of 1% agarose containing DPBS were placed on top of the samples in the dishes to immobilize bacteria and cells. Samples were imaged immediately using microscopy.

Conditioned Medium from Neutrophils

Approximately 2×106/mL of freshly harvested neutrophils were incubated at 37 °C in a roller drum spinning at a speed of 4 rotations per minute in PBGT medium (10 mM phosphate buffer pH 7.4 (2.46 mM monobasic with 7.54 mM dibasic sodium phosphate) containing 140 mM sodium chloride, 10 mM potassium chloride, 0.5 mM magnesium chloride, 1 mM calcium chloride, 1 mg/mL glucose, and 5 mM taurine [24]). To activate neutrophils, cultures were supplemented with 100 ng/mL of PMA for 1 hour and were separated from the supernatant by centrifugation for 5 minutes at 4,600 x g and 25 °C. The supernatant was isolated and used as conditioned medium. P. aeruginosa cultures were incubated for 3 hours in conditioned medium at 37 °C before imaging.

Growth Curves

Growth curve experiments were performed using a Synergy HTX multi-mode plate reader and sterile, tissue-culture treated, clear bottom, black polystyrene 96-well microplates (Corning, Corning, NY) containing 200 μL of culture in each well. The temperature set point was 37 °C and preheated before beginning measurements. For experiments performed with stationary phase bacteria, overnight cultures of bacteria were grown for 18 hours at 37 °C in the roller drum at 200 rpm to saturation, diluted into minimal medium to an optical density at 600 nm (OD600) of 0.01 containing NaOCl, H2O2, HNO3 or no supplement. Plates were incubated at 37 °C with continuous orbital shaking with an amplitude of 3 mm at the frequency of 180 cycles per minute and measured for OD600 every 20 minutes.

Reverse Transcription Quantitative PCR (RT-qPCR)

Strains were cultured overnight, diluted 1:1000 into MinA medium, incubated in a shaker at 225 rpm at 37°C for 2.5 hours, treated with 1 µM NaOCl, 4 µM H2O2, or untreated for 30 minutes, concentrated using syringe filters, and centrifuged at 13,000 × g for 4 minutes to pellet P. aeruginosa. Pellets were either snap-frozen in liquid nitrogen or processed immediately for RNA extraction using the NucleoSpin RNA kit (Macherey-Nagel, Allentown, PA). cDNA was prepared using the High-Capacity cDNA Reverse Transcriptase Kit (ThermoFisher). Quantitative PCR was performed using the SsoAdvanced Universal SYBR Green Supermix (BioRad, Hercules, CA) using the C1000 Thermal Cycler with a CFX96 real-time detection system (Bio-Rad). froA transcripts were quantified using the primers 5’-TTTCCCTCGCTTCCTCCGTC-3’ and 5’-ACCTTCCTTGGCCTTCTCGG-3’, which target PA14_21570 or PA3284 in PA14 or PAO1, respectively. The transcript abundance for each sample was normalized by 5S rRNA abundance, which was determined using the primers 5’-TAGAGCGTTGGAACCACCTG-3’ and 5’-GAGACCCCACACTACCATCG-3’ [48], yielding normalized count thresholds (Cts). The average fold change in transcript abundance was determined by computing the ratio of the averaged normalized Cts and raising by the power of 2, as performed previously [49].

Murine Model of Corneal Infection

The expression of froA was assessed in a murine model of corneal infection as described previously[50,51]. Briefly, P. aeruginosa strain PAO1F was cultured to mid-exponential phase in modified MinA minimal medium containing 0.05% citrate, 0.1% casamino acids (Gibco) and 0.2% glucose as the carbon sources and resuspended in PBS (Gibco). Corneal epithelia of C57BL/6J mice aged 6-8 weeks (Jackson Laboratory, Bar Harbor, ME) were abraded with 3×5 mm scratches using a 25 g needle, and 2 µL of PBS containing 5×104 of P. aeruginosa strain PAO1F was applied topically. Comparable numbers of male and female mice were used in each experimental group. Mouse eyeballs were collected and homogenized in 1 mL PBS at 2 hours post infection or 20 hours post infection. The homogenate was centrifuged for 5 minutes at 150 x g to remove corneal materials. Supernatant was centrifuged for 4 minutes at 13,000 x g. Pellets were suspended in lysis solution (10 mM Tris-HCl, 1 mM EDTA pH 8.0, 0.5 mg/mL lysozyme, 1% SDS) [49]. RNA was prepared using the NucleoSpin RNA kit (Macherey-Nagel) and assessed for mRNA expression using RT-qPCR.

RNA-seq Library Preparation and Data Analysis

P. aeruginosa was cultured and RNA was prepared as described in the RT-qPCR section. RNA yield was measured using a Nanodrop 2000 (Thermo Fisher, Waltham, MA). Samples were depleted of ribosomal RNA using the NEBNext rRNA Depletion kit (New Enland Biolabs, Ipswich, MA), from which cDNA libraries were constructed using the NEBNext Ultra II Directional Library kit (NEB), which were sequenced by the UC Irvine Genomics Research & Technology Hub (Irvine, CA) using an Illumina NovaSeq X Plus (Illumina, San Diego, CA) using paired-end 150 bp reads at a depth of approximately 10 M read per sample. Raw reads were checked with fastQC [52] (version 0.21.1), and trimmed and filtered into paired and unpaired reads using Trimmomatic [53] (version 0.39) in Java (1.8.0) using the ‘PE’ setting. Paired and unpaired reads were aligned to the PA14 genome (NCBI accession NC_008463.1) using Bowtie2 [54] (version 2.5.4) using the default ‘sensitive’ mode, were counted using the featureCounts program in Subread [55] (version 2.0.8) with the ‘countReadPairs’ option enabled for paired-end reads, and were fit into a negative binomial model to compute fold-change in gene expression using DeSeq2 [56] (version 1.40.2). The statistical significance of the changes were computed using the Wald test in DeSeq2. Gene Ontology enrichment analysis for Fig. 1A was performed using GOEnrichment [57] (version 2.0.1). The msrQ (PA14_62100), msrP (PA14_62110) genes were annotated at Pseudomonas.com [21] as yedZ and yedY, respectively, but are referred to here with their updated names [58].

Statistical Analysis

Statistical analysis and figures were generated in R (The R Foundation, Vienna, Austria) (version 4) and RStudio (Posit Software, Boston, MA).

Data, Code and Materials Availability

Raw data for all figures that contain statistical analyses are available in the Supplementary Information Source Data file. RNA-seq data is available at the National Center for Biotechnology Information Gene Expression Omnibus (GEO) (accession number GSE290217, accessible during review period using the token: whcbicwsvnmjdqj). The custom Matlab scripts used to process and analyze fluorescence microscopy data are freely available at https://github.com/asirya/AVIassembleGUI.

eLife Assessment

This is an important study reporting a new phenotype for a gene cluster that has previously been associated with the responses of the Gram-negative opportunistic pathogen Pseudomonas aeruginosa to flow fluid. Expression of the froABCD gene cluster is induced by HOCl in vitro and by activated immune cells, which produce these types of reactive chlorine species. Overall, the evidence presented by the authors is solid; however, the mechanism of fro-induction by HOCl remains unclear, and the evidence in support of the authors' claims is descriptive, which needs to be improved. This study is of interest to infection biologists interested in mechanisms of bacterial pathogenicity.

Reviewer #1 (Public review):

Summary:

Foik et al. report that hypochlorous acid, a reactive chlorine species generated during host defense, activates the transcription of the froABCD in P. aeruginosa. This gene cluster had previously been associated with a potential role during the flow of fluids and appears to be regulated by the sigma factor FroR and its anti-sigma factor FroI. In the present study, the authors show that froABCD is expressed both in neutrophils and macrophages, which they claim is likely a result of HOCl but not H2O2 production. Fro expression is also induced in a murine model of corneal infection, which is characterized by immune cell invasion. Expression of the fro system can be quenched by several antioxidants, such as methionine, cysteine, and others. FroR-deficient cells that lack froABCD expression during HOCl stress appear more sensitive to the oxidant.

Strengths:

The authors provide a number of data supporting their claim that transcription of the froABCD system is induced by reactive chlorine species. This was shown by RNAseq, qRT-PCR, and through microscopy using a transcriptional reporter fusion. Likewise, elevated expression of froABCD was shown in vitro and in vivo, excluding potential in vitro artifacts. The manuscript, while mostly descriptive, is easy to follow, and the data were presented clearly.

Weaknesses:

(1) Lines 60-62: Some of the authors' conclusions are not supported by the data and thus appear unfounded. One example: "we determine that fro upregulation.....These data suggest a novel mechanism..." Their data do not show that MSR upregulation is a direct effect of FroABCD. Instead, it could be possible that the FroR sigma factor also controls the expression of msr genes, which would be independent of froABCD.

(2) The authors show increased fro transcription both in neutrophils and macrophages; however, the two types of immune cells differ quite dramatically with respect to myeloperoxidase activation and HOCl production. Neither has this been discussed nor considered here.

(3) With respect to the activation of fro expression upon challenge with conditioned media from stimulated neutrophils, does the conditioned media contain detectable amounts of HOCl? Do chloramines, which are byproducts of HOCl oxidation with amines, also stimulate expression?

(4) A better control to prove that this fro expression is indeed induced by HOCl in activated neutrophils would be to conduct the experiments in the presence of a myeloperoxidase inhibitor.

(5) The work was conducted with two different P. aeruginosa strains (i.e. AL143 and PAO1F). None of the figure legends provides details on which strain was used. For instance, in line 111, the authors refer to Figure S1B for data that I thought were done with PAO1F, while in 154, data were presented in the context of the infection model, which was conducted with the other strain.

(6) It would be good if immune cell recruitment at 2hrs and 20hrs PI could be quantified.

(7) The conclusions of Figure 4 are, in my opinion, weak (line 187-188; "It is possible that ....."). These antioxidants likely quench the low amounts of NaOCl directly. This would significantly reduce the NaOCl concentrations to a level that no longer activates expression of fro. There is no direct evidence provided that oxidized methionine induces fro expression. Do the authors postulate that this is free methionine, or could methionine and/or cysteine oxidation in FroR increase the binding affinity of the sigma factor to the promoter? Another possibility is that NaOCl deactivates the anti-sigma factor. None of these scenarios has been considered here.

(8) Line 184: The reaction constants of HOCl with Cys and Met are similar.

(9) Treatment with 16 uM NaOCl caused a growth arrest of ~15 hrs in the WT (Figure 5A), whereas no growth at all was recorded with 7.5 uM in Figure 3A.

(10) The concentration range of NaOCl causing fro expression is extremely narrow, while oxidative burst rapidly generates HOCl at much higher concentrations. This should be discussed in more detail.

Reviewer #2 (Public review):

Summary:

Foik et al. studied the regulation of the fro operon in response to HOCl, an oxidant derived from immune cells, especially neutrophils. They use a transcriptional fusion of YFP to the froA promoter in an mCherry-expressing P. aeruginosa strain to determine fro-induction under the microscope. They use this system to study fro expression in medium, in the presence of neutrophils and macrophages, neutrophil-conditioned medium, and several chemical stimuli, including NaCl, HOCl, hydrogen peroxide, nitric acid, hydrochloric acid, and sodium hydroxide. They also use a corneal infection model to demonstrate that froA is upregulated in P. aeruginosa 20 h post-infection and perform transcriptional analyses in WT and a froR mutant in response to HOCl.

Strengths:

Their data clearly shows that HOCl is a strong inducer of the fro Operon. The addition of HOCl-quenching chemicals together with HOCl abrogates the response. They also show that a froR mutant is more susceptible to HOCl than WT. Their transcriptomic data reveal genes under control of the FroR/FroI sigma factor/anti sigma factor system.

Weaknesses:

Although the presented evidence is mostly solid, some of their findings need to be evaluated more carefully; explaining the rationale behind some of the experiments might enhance the article, and some of the models proposed by the authors seem far-fetched, as outlined below:

(1) In line 76 the authors claim "Relative to P. aeruginosa that were incubated in host cell-free media, P. aeruginosa in close proximity to human neutrophils or that were engulfed in mouse macrophages appeared to increase fro expression (Fig. 1C)". Counting bacterial cells in Figure 1C shows that 1 in 17 bacteria (5.8%) induce the froA-promotor in media in the absence of immune cells, while 4 in 72 bacteria (only 5.5%) do the same in the presence of neutrophils. Contrary to the authors' claims, it appears that P. aeruginosa actually decreases fro-expression in close proximity to neutrophils. There is a slight increase in fro-expression in bacteria co-incubated with macrophages (3 in 21, or 14.3%). A more rigorous statistical analysis might substantiate the authors' claim, but, as is, the claim "neutrophils increase fro expression" is untenable.

(2) The authors should explain the rationale behind some of the chemicals used. Why did they use nitric acid? Especially at these high concentrations, a strong acid such as nitric acid might have a significant influence on the medium pH. I understand that the medium is phosphate-buffered, but 25 mM nitric acid in an unbuffered medium would shift the pH well below 2. Similar considerations apply to hydrochloric acid and sodium hydroxide.

(3) In line 187, the authors state that "It is possible that oxidized methionine increases fro expression" and they suggest a model to that effect in Figure 5D. It is unclear why the authors singled out methionine sulfoxide, since a number of other things get oxidized by HOCl. In line 184, the authors state, in the same vein, that "HOCl oxidizes methionine residues 100-fold more rapidly than other cellular components". The authors should state which other cellular compounds they are referring to. Certainly not cysteine and other thiols, which react equally fast and are highly abundant in the cell: P. aeruginosa contains 340 µM GSH, 140 µM CoA-SH (https://doi.org/10.1074/jbc.RA119.009934) plus free cysteine and cysteines in proteins (based on codon usage, 1.34% of amino acids in proteins are cysteine, while methionine is only slightly more present at 2.10%, although a number of starting methionines are removed from mature proteins).

(4) Overall (and this is probably not addressable with the authors' data), some very interesting questions remain unanswered: what is the molecular mechanism of fro-induction? How is the FroR/FroI system modulated by HOCl? Does the system sense free or protein-bound methionine-sulfoxide? Are certain methionine residues in these proteins directly oxidized by HOCl? Many "HOCl-sensing" proteins are also modified at cysteine residues or amino groups; could those play a role? And lastly: what is the connection between shear/fluid flow and HOCl, or are these totally separate mechanisms of fro-induction?

Author response:

The following is the authors’ response to the original reviews.

We greatly appreciate the efforts of the reviewers, which have provided insightful and helpful comments to improve the manuscript. The feedback touches upon a number of topics, focusing on clarification or justification of experimental techniques and on understanding the mechanism by which P. aeruginosa detects HOCl. All reviewers raised the issue of how HOCl activates fro expression, including whether free or protein-bound methionine, cysteine, or other HOCl byproducts induce this expression. For the upcoming revision, we plan to perform experiments that address this issue and will discuss potential mechanistic models in light of the new data. In addition, we plan to perform additional experiments to address a reviewer’s concerns regarding the dependence of the fro response on HOCl production by neutrophils. The revision will correct imprecise statements pointed out by reviewers, and address all remaining issues requiring clarification or further discussion, including the range of HOCl sensitivity, relationship between HOCl and flow sensitivity, and justification for testing the fro response to nitric acid.

We have completed a number of experiments and responded thoroughly to reviewer comments below. We thank the reviewers again for their details comments, which suggested additional experiments and interpretations that have resulted in significant additional insight into the potential mechanism of HOCl sensing and its relevance with neutrophils.

Reviewer #1 (Public review):

Summary:

Foik et al. report that hypochlorous acid, a reactive chlorine species generated during host defense, activates the transcription of the froABCD in P. aeruginosa. This gene cluster had previously been associated with a potential role during the flow of fluids and appears to be regulated by the sigma factor FroR and its antisigma factor FroI. In the present study, the authors show that froABCD is expressed both in neutrophils and macrophages, which they claim is likely a result of HOCl but not H2O2 production. Fro expression is also induced in a murine model of corneal infection, which is characterized by immune cell invasion. Expression of the fro system can be quenched by several antioxidants, such as methionine, cysteine, and others. FroR-deficient cells that lack froABCD expression during HOCl stress appear more sensitive to the oxidant.

Strengths:

The authors provide a number of data supporting their claim that transcription of the froABCD system is induced by reactive chlorine species. This was shown by RNAseq, qRT-PCR, and through microscopy using a transcriptional reporter fusion. Likewise, elevated expression of froABCD was shown in vitro and in vivo, excluding potential in vitro artifacts. The manuscript, while mostly descriptive, is easy to follow, and the data were presented clearly.

We greatly appreciate the efforts of the reviewer and thank them for their succinct summary of the manuscript.

Weaknesses:

(1) Lines 60-62: Some of the authors' conclusions are not supported by the data and thus appear unfounded. One example: "we determine that fro upregulation.....These data suggest a novel mechanism..." Their data do not show that MSR upregulation is a direct effect of FroABCD. Instead, it could be possible that the FroR sigma factor also controls the expression of msr genes, which would be independent of froABCD.

We thank the reviewer for pointing out this important distinction. We have clarified in lines 63-65 in the clean version of the revision that MSR upregulation depends on FroR rather than FroABCD.

(2) The authors show increased fro transcription both in neutrophils and macrophages; however, the two types of immune cells differ quite dramatically with respect to myeloperoxidase activation and HOCl production.

Neither has this been discussed nor considered here.

We agree that the distinction between the cell types is important and have added a brief description of the differences in respiratory bursts and ROS production between the two cell types and our justification for focusing on neutrophils in lines 102-103. We think it’s very interesting that Fro appears to be activated by macrophages, which are not associated with HOCl production on their own. We think that it would be interesting to identify what is inducing Fro in macrophages in future work.

(3) With respect to the activation of fro expression upon challenge with conditioned media from stimulated neutrophils, does the conditioned media contain detectable amounts of HOCl? Do chloramines, which are byproducts of HOCl oxidation with amines, also stimulate expression?

This is an excellent question that addresses which molecules Fro is responding to from neutrophils. We have performed additional experiments that confirm that PMA-stimulated neutrophils produce HOCl (Fig. S2) through the use of a commercial hypochlorite sensor assay, which claims high specificity for detecting HOCl. We further confirmed that this production is inhibited by pretreatment with the MPO-specific inhibitor 4ABAH. These data support the interpretation that the Fro response to stimulated neutrophils requires MPO activity, of which the major product is HOCl. We have described this in lines 117-127.

Our data does not exclude the possibility that other MPO products could activate Fro expression. We were unable to obtain a reliable source of the major secondary MPO product, taurine chloramine, for our experiments, unfortunately. We believe that understanding the potential for secondary products to activate the response is an important and interesting question that can be explored in a future study. We have added a discussion of this in lines 367-376.

(4) A better control to prove that this fro expression is indeed induced by HOCl in activated neutrophils would be to conduct the experiments in the presence of a myeloperoxidase inhibitor.

We thank the reviewer for raising this point. We have performed the suggested set of experiments and found that indeed, the pre-treatment of neutrophils with MPO inhibitor 4-ABAH prior to PMA stimulation suppresses the activation of fro (Fig. 2D and Figure 2-figure supplement 1-2). The results are discussed in lines 117-127.

(5) The work was conducted with two different P. aeruginosa strains (i.e. AL143 and PAO1F). None of the figure legends provides details on which strain was used. For instance, in line 111, the authors refer to Figure S1B for data that I thought were done with PAO1F, while in 154, data were presented in the context of the infection model, which was conducted with the other strain.

We thank the reviewer for pointing this issue out. We have ensured that strain names appear in all the revised figure legends. To clarify, only mouse experiments and a related RT-qPCR assay used strain PAO1F due to prior IACUC approval of this strain and its use in previous publications.

(6) It would be good if immune cell recruitment at 2hrs and 20hrs PI could be quantified.

We previously quantified neutrophil recruitment at the site of corneal abrasion at 24 hours using the same conditions and strains (Ratitong, B. et al., J. Immun, 2022). While we do not have immune cell recruitment data for the 2 hr and 20 hr time points, the previous data show significant neutrophil recruitment near the latter time point, which is consistent with the interpretation that fro expression is activated by stimulated neutrophils. We have discussed this in lines 212-216.

(7) The conclusions of Figure 4 are, in my opinion, weak (line 187-188; "It is possible that ....."). These antioxidants likely quench the low amounts of NaOCl directly. This would significantly reduce the NaOCl concentrations to a level that no longer activates expression of fro. There is no direct evidence provided that oxidized methionine induces fro expression. Do the authors postulate that this is free methionine, or could methionine and/or cysteine oxidation in FroR increase the binding affinity of the sigma factor to the promoter? Another possibility is that NaOCl deactivates the anti-sigma factor. None of these scenarios has been considered here.

We acknowledge that our model of HOCl sensing was unclear and thank the reviewer for their insight. This critique is echoed by reviewer #2 in comment 3 as well. We recently found that the FroI anti-sigma factor has the highest concentration of methionine and cysteine residues of all known P. aeruginosa anti-sigma factors (Appendix 2—Table 1). Given that FroR and FroI form an extracytoplasmic function sigma – anti-sigma pair, which are associated with transducing extracellular signals to the cytoplasm, we have proposed an alternative model in which HOCl or secondary RCS molecules are detected through their oxidation of cysteine and methionine residues in FroI. This is discussed in lines 271-279 and lines 363-367.

(8) Line 184: The reaction constants of HOCl with Cys and Met are similar.

We thank the reviewer for pointing out this important clarification. We have revised the sentence to accurately reflect this in lines 243-244.

(9) Treatment with 16 uM NaOCl caused a growth arrest of ~15 hrs in the WT (Figure 5A), whereas no growth at all was recorded with 7.5 uM in Figure 3A.

We thank the reviewer for catching this. We have determined that the concentration of NaOCl in the reagent used for this particular experiment was lower than expected, thus requiring a higher concentration to achieve growth inhibition. We have repeated the experiment with new reagent and find that the results (now Figure 6A) are similar to the previous experiment but at a lower concentration of 4 micromolar, consistent with the concentration found to be sub-inhibitory in Figure 3A.

(10) The concentration range of NaOCl causing fro expression is extremely narrow, while oxidative burst rapidly generates HOCl at much higher concentrations. This should be discussed in more detail.

We appreciate the reviewer’s comment, which is related to reviewer #2’s comment #9. We have clarified the reported production rates of HOCl, which far surpass the bacterial MIC. After greater consideration, we believe secondary HOCl products including taurine chloramine could have a more significant role in vivo. While this molecule is less potent than HOCl, it is longer-lived, retains bactericidal activity, and retains the ability to oxidize methionine. We have discussed this in lines 377-405.

Reviewer #1 (Recommendations for the authors):

(1) Some statements in the text don't match the data shown in the Figures. For instance:

(a) Figure 2B shows ~65-fold fro expression, but the text states: "...increased expression of fro expression by 30fold..."(line 88).

The YFP/mCherry value of PMA-stimulated is 67.4 in the Figure (now Figure 2C). The fold-change is computed relative to unstimulated conditioned medium (third column, which has a value of 2.2), which is a 30-fold change. We have added a citation in the main text to the Source Data, which provides these raw values, to help clarify the computation for readers, and added in the legend that the value for unstimulated is greater than 1.

(b) Figure 3B shows ~35-fold fro expression at 1 uM NaOCl, but the text states: "...NaOCl increased fro expression by up to 74-fold..."(line 110).

We have clarified that the increase is relative to untreated, added that untreated value is below 1 in the caption, and provided a citation in the main text to the Source Data, which contains the raw values. The change is measured relative to untreated, for which the YFP/mCherry value is 0.48. The value at 1 uM is 35.7, giving a 74-fold change.

(2) Line 229: While the ∆froR strain was sensitive to HOCl, the strain was tolerant". Please revise.

We have corrected this typo (now lines 328- 329). We meant to convey that growth was not entirely inhibited in the froR strain.

Reviewer #2 (Public review):

Summary:

Foik et al. studied the regulation of the fro operon in response to HOCl, an oxidant derived from immune cells, especially neutrophils. They use a transcriptional fusion of YFP to the froA promoter in an mCherry-expressing P. aeruginosa strain to determine fro-induction under the microscope. They use this system to study fro expression in medium, in the presence of neutrophils and macrophages, neutrophil-conditioned medium, and several chemical stimuli, including NaCl, HOCl, hydrogen peroxide, nitric acid, hydrochloric acid, and sodium hydroxide. They also use a corneal infection model to demonstrate that froA is upregulated in P. aeruginosa 20 h post-infection and perform transcriptional analyses in WT and a froR mutant in response to HOCl.

Strengths:

Their data clearly shows that HOCl is a strong inducer of the fro Operon. The addition of HOClquenching chemicals together with HOCl abrogates the response. They also show that a froR mutant is more susceptible to HOCl than WT. Their transcriptomic data reveal genes under control of the FroR/FroI sigma factor/anti sigma factor system.

Weaknesses:

Although the presented evidence is mostly solid, some of their findings need to be evaluated more carefully; explaining the rationale behind some of the experiments might enhance the article, and some of the models proposed by the authors seem far-fetched, as outlined below:

We greatly appreciate the reviewer’s efforts and thank them for highlighting strengths and areas for improvement.

(1) In line 76 the authors claim "Relative to P. aeruginosa that were incubated in host cell-free media, P. aeruginosa in close proximity to human neutrophils or that were engulfed in mouse macrophages appeared to increase fro expression (Fig. 1C)". Counting bacterial cells in Figure 1C shows that 1 in 17 bacteria (5.8%) induce the froA-promotor in media in the absence of immune cells, while 4 in 72 bacteria (only 5.5%) do the same in the presence of neutrophils. Contrary to the authors' claims, it appears that P. aeruginosa actually decreases fro-expression in close proximity to neutrophils. There is a slight increase in fro-expression in bacteria co-incubated with macrophages (3 in 21, or 14.3%). A more rigorous statistical analysis might substantiate the authors' claim, but, as is, the claim "neutrophils increase fro expression" is untenable.

We believe the images alone do not give an adequate representation of the data and have quantified a larger portion of the data, which has been added as Figure 1D. The quantification supports the original claim that fro expression is increased during co-incubation with macrophages and neutrophils. Since there was not sufficient statistical sampling to distinguish engulfed P. aeruginosa from free ones, this part of the claim has been removed from the text (updated in lines 80-84).

(2) The authors should explain the rationale behind some of the chemicals used. Why did they use nitric acid? Especially at these high concentrations, a strong acid such as nitric acid might have a significant influence on the medium pH. I understand that the medium is phosphate-buffered, but 25 mM nitric acid in an unbuffered medium would shift the pH well below 2. Similar considerations apply to hydrochloric acid and sodium hydroxide.

We thank the reviewers for pointing out the need for this clarification. We have updated Figure 3D with a lower concentration of NaOH at 1 uM, which is the same concentration as NaOCl that activates fro expression. Due to the high buffering capacity of our medium, a high concentration of 6 mM NaOH was needed to induce a discernible change in pH and this high concentration of NaOH had no obvious effect on growth. Neither 1 uM nor 6 mM NaOH produced a change in fro expression, consistent with our previous findings that the effect is not due to sodium ions or higher pH. These updated findings are described in lines 181-190.

Since the effect of chloride is already controlled for using NaCl and the concentration of HCl used was not sufficient to cause a significant change in pH in the buffered medium, we have removed the HCl group from the data.

We have clarified that nitric acid was used because it is a strong oxidizer that is not found in neutrophils and that concentrations used were near the minimal inhibitory concentrations (lines 145-147 and lines 191-198). We acknowledge that the growth inhibition from HNO3 could be due to pH or oxidation. However, since no change in fro expression was observed at concentrations approaching the inhibitory concentration, we did not address the potential effects of low pH from nitric acid on fro expression.

(3) In line 187, the authors state that "It is possible that oxidized methionine increases fro expression" and they suggest a model to that effect in Figure 5D. It is unclear why the authors singled out methionine sulfoxide, since a number of other things get oxidized by HOCl. In line 184, the authors state, in the same vein, that "HOCl oxidizes methionine residues 100-fold more rapidly than other cellular components". The authors should state which other cellular compounds they are referring to. Certainly not cysteine and other thiols, which react equally fast and are highly abundant in the cell: P. aeruginosa contains 340 µM GSH, 140 µM CoA-SH (https://doi.org/10.1074/jbc.RA119.009934) plus free cysteine and cysteines in proteins (based on codon usage, 1.34% of amino acids in proteins are cysteine, while methionine is only slightly more present at 2.10%, although a number of starting methionines are removed from mature proteins).

We acknowledge that our HOCl sensing model had been vague and unclear and thank the reviewer for their insight. This critique is echoed by reviewer #1 in comment 7 as well.

Our initial suggestion that methionine sulfoxide was sensed was motivated by the observation that methionine sulfoxide reductases are upregulated by HOCl. However, we have revised this based on feedback from reviewers and further consideration of chlorine redox chemistry. Interestingly, we found that the FroI anti-sigma factor has the highest concentration of methionine and cysteine residues of all the P. aeruginosa anti-sigma factors (Appendix 2—table 1). Given that FroR and FroI form an extracytoplasmic function sigma – anti-sigma pair, which are associated with transducing extracellular signals to the cytoplasm, we propose a model in which HOCl or secondary RCS molecules are detected by their oxidation of cysteine and methionine residues in FroI. This is discussed in lines 271-278 and lines 362-366.

(4) Overall (and this is probably not addressable with the authors' data), some very interesting questions remain unanswered: what is the molecular mechanism of fro-induction? How is the FroR/FroI system modulated by HOCl? Does the system sense free or protein-bound methionine-sulfoxide? Are certain methionine residues in these proteins directly oxidized by HOCl? Many "HOCl-sensing" proteins are also modified at cysteine residues or amino groups; could those play a role? And lastly: what is the connection between shear/fluid flow and HOCl, or are these totally separate mechanisms of fro-induction?

We thank the reviewer for raising these excellent mechanistic questions. Issues relating to HOCl sensing are addressed in the preceding comment.

Regarding the connection to shear sensing, Padron et al., 2023 found that the detection of flow in P. aeruginosa can be attributed to chemical transport, in particular to H2O2 that was present in growth media. Based on the same principle, we expect Fro to be upregulated in flow at much lower concentrations than those observed in stationary fluids. The activation of Fro and the effects of HOCl would thus be expected to be flow-sensitive. We have commented on this important factor in the discussion in lines 406-417.

Reviewer #2 (Recommendations for the authors):

(1) To address 1, the authors could evaluate the microscopic images in the same manner in which they evaluated the other microscopic images, as, for example, presented in Figure 2B or Figure S1B.

See response to Weakness point (1).

(2) To address 2, please explain the choice of the chemicals (why nitric acid?), but also provide the pH of the media with those high concentrations of strong acids and bases, and interpret them in light of the permissible pH range for P. aeruginosa growth. More sensible controls might be a lower NaOH concentration in the range that would be reached through the amount of NaOH in the NaOCl stock at the highest NaOCl concentrations used. As for the acids, I don't see a reason to use these acids at these high concentrations. Please explain.

See response to Weakness point (2).

(3) To address 3: The authors could specify their methionine-sulfoxide model a bit more, so that testable hypotheses can be developed. If the authors think the FroR/FroI system senses free methionine sulfoxide, they or others could add methionine sulfoxide to the medium and check induction. If they think specific methionine residues in these proteins are oxidized, they could provide evolutionary evidence of conserved methionine residues. Or, based on a structure or structural prediction, they (or others) could mutate methionine residues, e.g., at the protein's surface or potential protein/protein-interaction sites and assess the effect on HOCl-based activation. Or they could consider other amino acids known to be highly reactive towards HOCl and mutate those in a future study.

See response to Weakness point (3).

Further comments:

(4) The headline of the figure legend of Figure S1 seems incomplete. Please mention the flow experiments shown in Figure S1A.

We have updated the title of this figure, which now appears in the eLife format as Figure 1 – figure supplement 1.

(5) What is the difference between the data presented in Figure 4C (bars "UTR" and "NaOCl") and the same bars in Figure S2A? Is this redundant or a re-plot?

In this revision, Figure 4C has become Figure 5B and Figure S2A is now Figure 5—figure supplement 1. Only the 1 uM NaOCl condition is replotted. We have described this in the legend for Figure 5—figure supplement 1.

(6) Line 228: hpd is more likely a gene of the aromatic amino acid catabolism.

We thank the reviewer for pointing this out. We have removed the ‘branched’ descriptor in this sentence, now in lines 325-326.

(7) Line 229: "While the ΔfroR strain was sensitive to HOCl, the strain was tolerant (Figure 5A)". Please clarify. Which strain was tolerant? WT?

We have corrected this typo (now line 328-329). We meant to convey that growth was not entirely inhibited in the froR strain.

(8) Line 247: "Activated neutrophils produce HOCl concentrations as high as 50 µM [24]." This "50 µM" number is often quoted; the citation trail typically leads to Weiss et al. 1982 (https://doi.org/10.1172/JCI110652). However, a more factually correct statement based on that paper would be "2 x 10^6 neutrophils, activated with 30 ng/mL PMA at 37C in 1 mL of Dulbecco's buffer can produce around 50 nmol HOCl per hour". In the particular reference 24, Dybpukt et al used a methodology similar to Weiss et al., and here around 50nmol were produced by the same number of cells in 30 min in response to 100 ng/mL PMA. Please clarify accordingly.

We thank the reviewer for bringing this to our attention. We have altered the language to indicate that the production rate is for this specific set of parameters. Related to this, Reviewer #1 Comment #10 requested a more detailed discussion of the significance of the Fro response, since it is at much lower HOCl concentration than produced by neutrophils. We have discussed this in lines 377-405.

(9) Line 275: "which is strain PA14 strain". Please clarify.

We have fixed this error and entered it into the Key Resource Table.

(10) Line 311: "Cultures containing densities below 10 P. aeruginosa per frame were concentrated using a syringe filter with 0.2 or 0.8 µm pore sizes (Millipore, Burlington, MA)." Isn't that a bit concerning when testing the induction of an operon that is supposedly activated by shear through fluid flow? Did the authors convince themselves that this procedure does not induce fro?

We do not expect the filtering procedure to cause changes in gene expression because cells are imaged immediately after filtering. Nonetheless, we performed additional experiments (Fig 3F and 2D) entirely without concentrating cells and found that YFP/mCherry levels were consistent with previous data in Figure 3. This rationale has been added to the Methods section under the “Fluorescence and Phase Contrast Microscopy” section.

Version 2

Abstract

Neutrophils kill bacteria by producing an array of lethal molecules, including hypochlorous acid (HOCl). The extent to which pathogens detect HOCl from neutrophils and defend themselves appropriately has not been understood. We report that the opportunistic pathogen Pseudomonas aeruginosa responds to activated neutrophils by upregulating its expression of an operon which was previously shown to be activated by fluid flow, and named the flow regulated operon (Fro). We found that fro is upregulated in a mouse infection model where neutrophil influx occurs, specifically by HOCl and not by other neutrophil defense factors including LL-37, histones, or H2O2. HOCl preferentially oxidizes methionine and cysteine residues. As the Fro operon is regulated by the anti-σ factor FroI, which contains the highest density of these residues of all anti-σ factors, this raises the possibility that FroI is an HOCl sensor. In support of this, we found that via the σ factor FroR, HOCl induces upregulation of methionine sulfoxide reductases that relieve otherwise lethal oxidative stress. These findings suggest a model in which flow transports oxidizing molecules that activate the fro operon, establishing an early warning system for P. aeruginosa that improves its survival against host immune defenses and persistence during infection.

Introduction

The Gram-negative opportunistic pathogen P. aeruginosa is a major cause of hospital-acquired infections and is a significant antibiotic resistance threat. To survive in a broad range of conditions, this bacterium has evolved a variety of signaling pathways that detect changes in nutrients, chemical gradients, and flow, and activates responses that promote its survival (Fang et al., 2016; Sanfilippo et al., 2019). Understanding how extracellular-sensing pathways detect and respond to host cues is critical for improving the treatment of bacterial infections and minimizing the generation of antibiotic-resistant bacteria.

Mammalian immune systems respond to the presence of bacterial pathogens by recruiting immune cells that migrate rapidly through the circulatory system to sites of infection (Aymonnier et al., 2023). Neutrophils are the first and most abundant immune cells recruited to inflamed or damaged tissues (Lavoie et al., 2011) and are stimulated into an “activated” state by bacterial components, such as the membrane constituent lipopolysaccharide. Activated neutrophils produce multiple products that are toxic to bacteria including reactive oxygen species (ROS) and reactive chlorine species (RCS) (Hampton et al., 1998; Segal, 2005; Sultana et al., 2020; Ulfig and Leichert, 2020).

While the toxic effects of ROS on bacteria are understood, far less is known about how bacteria are affected by RCS (Fang, 2011; Gray et al., 2013). Hypochlorous acid (HOCl) is a major RCS product of neutrophils that is formed through myeloperoxidase enzymes from hydrogen peroxide (H2O2) and chloride ions, and is strongly bactericidal (Albrich and Hurst, 1982; Harrison and Schultz, 1976; Segal, 2005; Winterbourn et al., 2016). P. aeruginosa is a potent activator of neutrophil respiratory bursts (Jensen et al., 1990) but whether and how the bacteria detect and respond to neutrophil components is not understood.

P. aeruginosa responds to shear generated by the flow of fluids via the fro operon, which is upregulated in the bacterium during infection in humans (Cornforth et al., 2018; Padron et al., 2023; Sanfilippo et al., 2019) (Figure 1A). In addition, fro is necessary for successful colonization of the gastrointestinal tract and lung infections (Potvin et al., 2003; Skurnik et al., 2013), although the reasons for this is unclear. Given its important role in infection, it is possible that Fro could respond to immune-associated cues, such as the presence of neutrophils. The operon consists of PA14_21570, PA14_21580, PA14_21590, and PA14_21600, which was termed froABCD (Sanfilippo et al., 2019), and the extracytoplasmic function (ECF) sigma (σ) and anti-σ factors PA14_21550 and PA14_21560 (Boechat et al., 2013), which were termed froR and froI respectively, by (Sanfilippo et al., 2019). Expression of froABCD is regulated by FroR and FroI: overexpression of froR upregulates fro, whereas overexpression of froI downregulates fro (Sanfilippo et al., 2019). Flow was reported to activate FroR-dependent transcription of froABCD through a mechanism that involves chemical transport (Padron et al., 2023).

Activation of fro promoter transcription by flow and during infection of human tissue.

(A) RNA-seq heatmap of P. aeruginosa genes that are upregulated by at least 3-fold by flow and during infection in human wounds. Datasets are from (Cornforth et al., 2018; Sanfilippo et al., 2019). Log2(FC) indicates the log2 of the fold-change. (B) Schematic of the P. aeruginosa fluorescent reporter strain AL143 that contains a transcriptional fusion of yfp to the fro promoter and expresses mCherry under a constitutive promoter. (C) Representative phase-contrast and fluorescence images and (D) corresponding fro expression as determined by the ratio of YFP to mCherry fluorescence, of P. aeruginosa strain AL143 co-incubated with neutrophils, macrophages (MΦs), or host cell-free medium for 30 minutes. Dashed lines in images indicate boundaries of host cells and white arrows indicate P. aeruginosa with intense YFP fluorescence relative to that of mCherry. Data points indicate individual P. aeruginosa cells from a single experiment and bars indicate the average. Images in (C) are representative of three independent experiments and scale bars represent 5 μm. P-values were obtained using Welch’s t-test of data from single experiments, with values of p>0.05 denoted as ns.

Here, we show that the fro operon responds directly to HOCl, which is produced robustly by neutrophils during immune response. Using a reporter system in live bacteria and through quantitative measurements of transcription, we show that fro expression is upregulated by activated neutrophils and in a murine model of corneal infection. Through transcriptional profiling, we determine that bacteria mitigate the toxic effects of HOCl through a FroR-dependent mechanism that activates methionine sulfoxide reductases to relieve oxidative stress. These data suggest a novel mechanism in which bacterial defense against host innate immunity is primed by a single system that detects both flow and HOCl.

Results

Macrophages and neutrophils activate fro expression in P. aeruginosa under non-flow conditions

To investigate whether immune cells directly activate the fro operon, we quantified fro expression using a P. aeruginosa strain that co-expresses the yellow fluorescent protein (YFP) under the transcriptional control of the fro promoter and mCherry under the control of a constitutive promoter (Figure 1B). While fro expression increased under flow (Figure 1—figure supplement 1A), experiments in this study were performed in the absence of flow in order to deconvolve the effects of flow from potential activation of fro by immune cells. Because neutrophils and macrophages are the first immune cells recruited to sites of infection, we co-incubated P. aeruginosa with human neutrophils or mouse macrophages for 30 minutes. Relative to P. aeruginosa that were incubated in host cell-free media, P. aeruginosa co-incubated with either neutrophils or macrophages increased YFP/mCherry ratios by comparable levels (Figure 1C-D), suggesting that close proximity to or direct interaction with these host immune cells increases fro expression in P. aeruginosa.

Activated neutrophils induce fro expression

To assess whether fro upregulation in P. aeruginosa could be part of a response to respiratory bursts from neutrophils and macrophages, we focused on the fro response to neutrophils (Figure 2A) because these cells produce larger respiratory bursts and greater levels of ROS than macrophages (Colombo et al., 2017; Forman and Torres, 2002; Thekkan et al., 2019). P. aeruginosa were incubated in conditioned medium from human neutrophils that were stimulated using phorbol 12-myristate 13-acetate (PMA), which triggers respiratory bursts (Karlsson et al., 2000). The expression of fro was quantified by averaging the ratios of YFP to mCherry fluorescence for at least one hundred individual P. aeruginosa cells. In support of our hypothesis, conditioned medium from PMA-stimulated neutrophils increased fro expression in bacteria by 30-fold compared to conditioned medium from unstimulated neutrophils (Figure 2B-C, Figure 2—source data 2). We confirmed that this fro response was not due to PMA alone, as PMA-treated medium without neutrophils had no effect on fro expression (Figure 2B-C). Similarly, conditioned medium from unstimulated neutrophils had no significant effect on fro expression compared to medium alone, suggesting that the upregulation of fro is due to respiratory burst products.

Conditioned medium from stimulated neutrophils induces fro expression.

(A) Schematic depicting that PMA-stimulated neutrophils produce respiratory burst products that could activate fro expression. (B) Representative phase-contrast and fluorescence images and (C) fro expression as determined by ratios of YFP to mCherry fluorescence, of P. aeruginosa strain AL143 cultured for 3 hours in medium only, medium with PMA, conditioned medium from PMA-stimulated neutrophils, or conditioned medium from unstimulated neutrophils. Unstimulated neutrophils have a value greater than 1. Scale bars represent 5 μm. Data points indicate an average from at least one hundred individual P. aeruginosa and gray columns indicate the mean of at least three independent experiments. (D) fro expression as determined by ratios of YFP to mCherry fluorescence of P. aeruginosa strain AL143 that was cultured for 3 hours in conditioned medium in which neutrophils were stimulated with 100 ng/mL PMA, pre-treated with 2mM 4-ABAH, or both. Data points indicate individual P. aeruginosa cells, with each color representing an independent experiment, bars represent the average of 3 independent experiments, each using separate neutrophil isolations, and error bars indicate standard error of the mean (SEM). P-values were obtained using Welch’s t-test, with values of p>0.05 denoted as ns.

HOCl and other RCS are produced from ROS during respiratory bursts via myeloperoxidase (MPO). To investigate whether P. aeruginosa fro expression is induced by MPO-derived RCS, we pre-treated neutrophils prior to PMA stimulation with 4-aminobenzoic acid hydrazide (4-ABAH) (Figure 2D), a specific and irreversible MPO inhibitor (Kettle et al., 1997). Indeed, the activation of fro expression by PMA-stimulated neutrophil medium was abrogated in neutrophils that were pre-treated with the MPO inhibitor. The degree of fro inhibition by 4-ABAH followed a dose-dependent trend (Figure 2—figure supplement 1), consistent with a graded fro response to different RCS concentrations. HOCl production in neutrophils due to PMA stimulation and its inhibition by 4-ABAH was confirmed using a fluorometric HOCl-specific assay (Figure 2—figure supplement 2). These results indicate that fro upregulation depends on MPO activity and suggest that fro is activated specifically in response to HOCl.

HOCl but neither H2O2 nor HNO3 induces fro expression

ROS and RCS have inhibitory effects on bacterial growth. To explore the hypothesis that these classes of molecules induce fro expression in P. aeruginosa, we measured the effects of H2O2, HNO3, and NaOCl on fro expression at concentrations approaching their minimum inhibitory concentrations (MICs) (Figure 3A). NaOCl increased fro expression 74-fold at 1 µM relative to untreated (Figure 3B, Figure 3B–Source Data). To validate that the increase in YFP intensity is due to increased transcription, reverse transcription quantitative PCR (RT-qPCR) was also performed, which yielded a 200-fold increase in fro transcription in response to 1 µM NaOCl treatment (Figure 3C). While 2 µM NaOCl caused only a 3-fold increase in fro expression, we attribute this to the effects of NaOCl on transcription and cell viability, as this concentration also decreased the constitutive mCherry reporter intensity and cell size (Figure 3—figure supplement 1).

NaOCl but neither H2O2 nor HNO3 induces fro expression.

(A) Growth profiles as measured by optical density (OD600) of P. aeruginosa treated with the indicated concentrations of NaOCl, H2O2, HNO3, or no treatment. Data points indicate a mean of at least two independent experiments and error bars indicate SEM. (B) fro expression as determined by ratios of YFP to mCherry fluorescence in P. aeruginosa after 3 hour treatment with the indicated concentrations of NaOCl or no treatment (UTR). The value for UTR is less than 1. (C) Abundance of froA transcripts in P. aeruginosa after 30 minutes of treatment with 1 µM NaOCl, 4 µM H2O2, or no treatment (UTR), as measured by RT-qPCR. Measurements were normalized to 5S ribosomal RNA and the logarithm (base 2) of the fold-change in transcription was computed relative to the untreated condition. Horizontal bars indicate the average of four independent experiments and error bars indicate SEM. (D and E) fro expression as determined by ratios of YFP to mCherry fluorescence in P. aeruginosa after 3 hour treatment with (D) 1 µM NaOCl, 1 µM NaCl, 1 µM NaOH, 1 µM or 6 mM NaOH, or no treatment, and (E) indicated concentrations of H2O2, HNO3, or no treatment (data replotted from panel B). In panels B-E, data points indicate the average from at least one hundred individual P. aeruginosa cells, horizontal bars indicate the mean of independent experiments and error bars indicate SEM. Dashed lines indicate the average of the untreated condition, which was normalized to one. (F) fro expression as determined by ratios of YFP to mCherry fluorescence of P. aeruginosa after 3 hour treatment with 0.5 µg/mL tobramycin, vehicle control (VC; sterile water), 40 µg/mL, 1 µM NaOCl, or LL-37 and 10 µg/mL histones (LL37+His). Data points indicate individual P. aeruginosa cells, with different colors representing independent experiments. Horizontal bars indicate the mean of independent experiments and error bars indicate SEM. P-values were obtained using Welch’s t-test, with values of p>0.05 denoted as ns. All data were acquired using the P. aeruginosa strain AL143.

To rule out that the individual sodium or chloride ion components of NaOCl were responsible for the increase in fro expression, we repeated experiments using NaCl, which yielded no change in fro expression (Figure 3D). To assess whether the fro response may be affected by pH, we tested NaOH at the same concentration as NaOCl, and at a concentration at which a change in pH could be discerned. Due to the robust buffering capacity of the medium, a pH change of ∼0.1 required 6 mM NaOH, which is a concentration that had no obvious effect on growth but is significantly higher than the concentration of NaOCl needed to activate fro expression. Regardless, neither concentration of NaOH had any significant impact on fro expression (Figure 3D). As fro activation is not due to sodium or chloride ions nor changes in pH, we attributed the effects to hypochlorite (OCl-), which is in equilibrium with hypochlorous acid (HOCl) at the pH of 7.1 in the medium used in our experiments.

To address the possibility that fro expression is activated by oxidation alone, we quantified fro expression in response to H2O2 and HNO3, at concentrations up to their respective MICs (Figure 3A). While both are strong oxidizers, HNO3 is not produced by neutrophils. Neither H2O2 nor HNO3 induced any significant changes in fro expression up to near-MIC concentrations (Figure 3E). The lack of fro expression due to H2O2 treatment is not due to defects in fluorescent protein production, as H2O2 did not induce any change in fro transcript abundance, as assayed using RT-qPCR (Figure 3C).

Together, these indicate that strong oxidation alone is not sufficient to upregulate fro. We assessed whether other antimicrobial components of neutrophils or other bacterial stressors could be responsible for the fro response. Stimulated neutrophils release the antimicrobial peptide LL-37 and histones, which together constitute a synergistic antimicrobial attack (Doolin et al., 2020; Duong et al., 2025). Tobramycin is an aminoglycoside, a class of antibiotics that induces oxidative stress within bacteria (Kohanski et al., 2008). Neither the combination of LL-37 and histones nor tobramycin at near-MIC concentrations induced fro expression (Figure 3F). Together, these results indicate that fro expression is activated specifically by the neutrophil effector HOCl and not by intracellular oxidative stress or other neutrophil defense factors including LL-37, histones, or H2O2.

Corneal infection activates fro expression in bacteria

P. aeruginosa is a major cause of corneal ulcers and keratitis, which can cause visual impairment and blindness. Neutrophils are recruited within hours to corneal epithelial abrasions, with neutrophil cell density peaking at 18 hours (Li et al., 2006). We tested the hypothesis that fro expression is upregulated during corneal infection using a murine bacterial keratitis model in which P. aeruginosa strain PAO1F is inoculated onto a corneal abrasion, which we previously demonstrated results in significant neutrophil recruitment to the cornea (Minns et al., 2023; Ratitong et al., 2022). Mice corneas were harvested 2 or 20 hours post-infection and fro transcripts were quantified using RT-qPCR. Transcript abundance was normalized to the number of P. aeruginosa cells using 5S ribosomal RNA. In support of the hypothesis, fro expression increased significantly by 20 hours post-infection (Figure 4A). Expression levels showed no significant change within the first two hours following infection, which we attribute to a lower level of immune cell recruitment during this period. The increase in fro expression 20 hours post-infection was not due to differences between the P. aeruginosa strains, because fro transcription in PAO1F also increased in response to NaOCl but not H2O2 (Figure 4—figure supplement 1), consistent with previous results.

The expression of fro is activated during corneal infection

Abundance of froA transcripts in P. aeruginosa strain PAO1F as measured by RT-qPCR, at 0, 2, or 20 hours post-inoculation onto abraded mouse corneas. Measurements were normalized to 5S ribosomal RNA and the logarithm (base 2) of the fold-change in transcription was computed relative to the initial inoculum (time=0 hrs). Log2(FC) indicates the log2 of the fold-change. Data points represent independent corneal infections. Horizontal bars indicate the average of at least six independent experiments and error bars represent SEM. P-values were obtained using Welch’s t-test, with values of p>0.05 denoted as ns.

Methionine sulfoxide reductase upregulation requires FroR

FroR is the σ factor required for activation of the fro operon in response to flow (Sanfilippo et al., 2019). To understand the response to HOCl, we performed transcriptional profiling of wild-type (WT) and ΔfroR P. aeruginosa strains. The genes that were most upregulated by NaOCl in the WT strain compared to the ΔfroR mutant were those in the fro operon, the phosphate and pyrophosphate-specific outer membrane porin genes oprO and oprP, and the methionine sulfoxide reductase genes msrB, msrP, and msrQ (Figure 5A and Appendix 1—table 1). Methionine sulfoxide reductases (MSRs) have a critical role in bacterial survival and defense against RCS stress. HOCl oxidizes methionine and cysteine residues 100-fold more rapidly than other amino acids (Pattison and Davies, 2006; Winterbourn, 1985) and oxidation of methionine in bacterial membrane proteins by neutrophil-produced HOCl correlates with increased bactericidal killing of E. coli and S. aureus (Rosen et al., 2009). HOCl oxidation of methionine forms methionine sulfoxide, which impairs protein function.

Expression of fro is inhibited by methionine and antioxidants.

(A) Heatmap showing P. aeruginosa gene transcripts (sorted by genomic locus) that were upregulated by at least 4-fold in the wild type PA14 background (strain AL143) compared to the ΔfroR mutant (AL325) after 30 minutes of treatment with 1 µM NaOCl and that had p-values less than 0.05 (n≥3). Log2(FC) indicates the log2 of the fold-change. Raw data can be found online (Foik, Ilona P. et al., 2025) in NCBI’s Gene Expression Omnibus (see Key Resources Table). (B and C) fro expression as determined by ratios of YFP to mCherry fluorescence in strain AL143 after 3 hour (B) treatment with; vehicle control (UTR), 1 µM NaOCl, or 1 µM NaOCl with combined with methionine (Met), cysteine (Cys), or β-mercaptoethanol (βME) at the indicated concentrations, and (C) incubation in medium only (replotted from Figure 2C), or conditioned medium from stimulated neutrophils with and without 100 µM methionine. Data points indicate an average from at least one hundred individual P. aeruginosa. Gray columns indicate the mean of at least three independent experiments and error bars represent SEM. P-values were obtained using Welch’s t-test, with values of p>0.05 denoted as ns.

MSRs restore proper protein function by reducing methionine sulfoxide back to methionine (Chandra et al., 2023; Gray et al., 2013). Our data thus suggest that FroR is required for the upregulation of MSRs in response to HOCl stress. We did not observe upregulation in PA14_27570, which is homologous to MsrC, which reduces strictly free methionine sulfoxide, or in PA14_66330 (MsrA), which reduces both free and protein-bound methionine sulfoxide (Appendix 1—table 1). No significant changes in the expression of enzymes that repair cysteine oxidation, such as thioredoxin, glutathione redoxin, or disulfide repair, were observed, suggesting that the fro response is directed towards repairing methionine oxidation.

We reasoned that HOCl could oxidize methionine residues within a P. aeruginosa sensor, which could in-turn upregulate fro expression. A number of anti-σ factors contain sensory domains that transduce sensory signals to their cytoplasmic σ factor (Paget, 2015). FroI is an anti-σ factor for the FroR σ factor (Boechat et al., 2013), which raises the possibility that HOCl or its secondary products could oxidize FroI directly, resulting in the activation of FroR. Consistent with this possibility, we found that FroI contains the highest concentration of methionine and cysteine residues of all identified P. aeruginosa PA14 anti-σ factors (Appendix 2—table 1).

We hypothesized that fro activation would be suppressed in the presence of excess of free methionine, as it would reduce available HOCl and thus oxidation of the FroI sensor. Indeed, co-treatment of NaOCl with methionine suppressed fro expression (Figure 5B). We reasoned that other antioxidants such as cysteine and β-mercaptoethanol (βME) could also function as alternative oxidation targets for HOCl, and therefore also suppress fro. Consistent with this hypothesis, co-treatment of NaOCl with cysteine or βME significantly suppressed fro expression compared to cysteine or βME alone (Figure 5B and Figure 5—figure supplement 1A-B).

We considered how fro could be upregulated by stimulated neutrophils in the context of these findings. HOCl and other reactive chlorine species produced by neutrophils could oxidize the P. aeruginosa sensor, resulting in fro upregulation. Under this interpretation, supplying excess methionine, which would reduce HOCl, should also inhibit activation by neutrophils. Indeed, supplementing conditioned media from PMA-stimulated neutrophils with methionine completely suppressed fro expression (Figure 5C). Together, these data suggest a model in which fro responds to HOCl-induced oxidation of a sensor by upregulating MSRs.

FroR protects P. aeruginosa against HOCl

The upregulation of MSRs is expected to improve bacterial growth against HOCl stress. We measured the growth profiles of WT and ΔfroR strains of P. aeruginosa. Treatment with 4 µM NaOCl significantly inhibited the growth of the wild-type strain (Figure 6A) but resulted in even greater growth inhibition in the ΔfroR strain. This observation suggests that the fro operon improves P. aeruginosa tolerance to HOCl stress.

FroR improves P. aeruginosa growth against HOCl stress and regulates transcription of genes involved in antioxidant defense.

(A) Growth profiles of wild-type (WT) P. aeruginosa strain AL143 and the ΔfroR mutant strain AL325 treated with 4 µM NaOCl or vehicle control (VC). Data points represent an average of at least three independent experiments and error bars indicate SEM. (B-C) Statistical significance as a function of fold-change in transcript abundance in (B) WT and (C) ΔfroR after treatment with 1 µM NaOCl (n≥3), as determined by RNA sequencing. Genes in red (upregulated) and blue (downregulated) were altered by at least four-fold. The unlabeled gene loci shown have not been previously annotated. P-values were determined using the Wald test (Love et al., 2014). Greater values along the vertical axis indicate greater statistical significance. P-values were determined using the Wald test (Love et al., 2014). Greater values along the vertical axis indicate greater statistical significance and the dashed horizontal line indicates a p-value of 0.05. Raw data can be found online (Foik, Ilona P. et al., 2025) in NCBI’s Gene Expression Omnibus (see Key Resources Table). (D) Schematic that depicts a model in which RCS oxidizes methionine residues (Met) in a target sensor(s), such as the ECF anti-σ factor FroI, which in turn activates the FroR σ factor and increases the production of methionine sulfoxide reductases (MSR).

We considered other genes which may improve WT P. aeruginosa tolerance to HOCl. HOCl increased the expression of iron uptake genes (pvdA and pvdQ) and downregulated the expression of oxidative phosphorylation genes (cyoCDE) (Figure 6B and Appendix 1—table 2). These changes suggest that HOCl causes an iron deficiency and decreases oxidative phosphorylation. In order to compensate for decreased oxidative phosphorylation, P. aeruginosa may utilize amino acid catabolism as an alternative energy source, as evidenced by a concomitant upregulation of amino acid catabolism genes (hpd, bkdA1, bkdA2, and bkdB) by HOCl (Figure 6B and Appendix 1—table 2).

While the ΔfroR mutant was more sensitive to HOCl than WT, its growth was not fully inhibited (Figure 6A), suggesting that additional mechanisms relieve HOCl stress. In the ΔfroR mutant, HOCl upregulated genes encoding multidrug efflux pumps (mexEF-oprN and mexXY) (Figure 6C and Appendix 1—table 3). These pumps can relieve oxidative stress (Fargier et al., 2012; Fraud and Poole, 2011) but were not activated in the WT strain. This observation suggests that bacterial defense against HOCl may be hierarchical, in which Fro-regulated and Mex pathways could be first-line and second-line responses to HOCl stress, respectively. Supporting this interpretation, it has been reported that treatment of the WT strain using significantly higher (millimolar) HOCl concentrations upregulates mexEF-oprN (Farrant et al., 2020) and the mexXY regulator mexT (Groitl et al., 2017).

Discussion

Neutrophils are activated by the presence of bacteria that infect host tissue. In their activated state, neutrophils generate respiratory bursts that produce an abundance of HOCl. We have demonstrated that P. aeruginosa responds to activated neutrophils by upregulating the transcriptional expression of the fro operon, which relieves HOCl stress. Our data suggest that the upregulation of fro could function as a bacterial defense mechanism against neutrophil attack, thus improving P. aeruginosa survival and colonization of human tissue during infection.

Detection and response to HOCl

Based on (i) our observation that the expression of multi-drug efflux pumps is activated in the absence of FroR (Figure 6C and Appendix 1—table 3), (ii) that this expression occurs at high HOCl concentrations (Farrant et al., 2020), and (iii) that multiple systems repair HOCl damage, the HOCl response could follow a hierarchy of activation. Our data suggest a model in which the fro system is the first-line response to HOCl. FroR activates two distinct classes of reductases that repair HOCl-oxidized methionine in proteins: MsrB reduces methionine sulfoxide in the cytoplasm using electrons provided by the thioredoxin system (Ezraty et al., 2005; Lourenço Dos Santos et al., 2018), while MsrPQ works in the periplasm and inner membrane using electrons from the respiratory chain (Gennaris et al., 2015). The activation of these two reductase types enables P. aeruginosa to limit HOCl stress in separate cellular spaces by utilizing distinct electron donor sources.

How does P. aeruginosa detect HOCl? As an anti-σ factor, FroI could immobilize FroR, inhibiting FroR from reaching transcriptional targets. Given the particularly high density of methionine and cysteine residues in the FroI, this protein could serve as an HOCl sensor by releasing the sigma factor FroR in response to oxidation of these residues (Figure 6D). While our data shows that HOCl can directly activate fro expression, HOCl reacts with neutrophil components, forming secondary RCS, including monochloramines, protein-derived chloramines, and taurine chloramine (Kim and Cha, 2014; Klebanoff et al., 2013; Ulfig and Leichert, 2020). Taurine chloramine is an expected major RCS released from neutrophils due to the high abundance of taurine in these cells (Kim and Cha, 2014; Marcinkiewicz and Kontny, 2014). Because it retains oxidizing potential towards cysteine and methionine (Peskin and Winterbourn, 2001), taurine chloramine could contribute to the activation of fro expression alongside HOCl. Future work will need to assess the relative efficacies of HOCl and other secondary RCS towards activating fro expression.

Relationship with neutrophil-mediated killing and flow

One of the striking features of the Fro system is that it is tuned for maximal activation at micromolar HOCl concentrations (1 µM) that are close to the MIC. While our experiments measured the effects of HOCl at steady-state using medium that does not contain amino acids, host proteins, and other host factors, HOCl dynamics in host environments are much more complex, where HOCl production is estimated to be in quantities that exceed the sensitivity of Fro. For example, HOCl production is predicted in excess of 100 mM per minute in the phagosome (Winterbourn et al., 2006). In addition, stimulated neutrophils at densities comparable to those found in blood (Tahir and Zahra, 2026) using PMA produces extracellular HOCl at 80 µM per hour, as measured through taurine chloramine absorbance (Weiss et al., 1982). Additional measurements using slight changes in stimulation time, neutrophil density, or PMA concentration yielded comparable HOCl production rates within a factor of 2 (Dypbukt et al., 2005; Test et al., 1984). At sites of inflammation, HOCl production in interstitial fluid has been predicted to be in the millimolar range due to the greater density of neutrophils there (Weiss, 1989).

Despite the high production rates of HOCl, the steady-state concentration in neutrophils and in vivo is expected to be significantly lower due to its rapid conversion into secondary RCS through reaction with neutrophil or plasma components. For example, the lifetime of HOCl is estimated to be 0.1 µs and have an action radius under 0.1 µm in the phagosome (Schürmann et al., 2017), which would mitigate the antimicrobial effects of HOCl but retain oxidizing potential. The secondary RCS taurine chloramine is long-lived, well-tolerated in the body, and retains significant bactericidal activity in the micromolar range (Gottardi and Nagl, 2010; Kim and Cha, 2014; Nagl et al., 2000). It is plausible that diminished HOCl or longer-lived secondary RCS such as taurine chloramine fall within the activating range of Fro, which would in turn mitigate their antimicrobial effects through upregulation of the MSRs. Importantly, our observation that fro expression is activated during corneal infection, which involves significant neutrophil recruitment, supports a role for the Fro response in a neutrophil-dense environment.

An additional layer of complexity of understanding the role of Fro during infection is that flow exposes bacteria to higher concentrations of molecules in a fluid. The bacterial response to flow, which was initially referred to as rheosensing (Sanfilippo et al., 2019), was more recently attributed to the effects of molecular transport by flow (Padron et al., 2023). In particular, flow stimulated a response to H2O2 in cells that were otherwise unresponsive to it in the absence of flow (Padron et al., 2023). Flow could similarly amplify the stimulating effects of RCS, causing P. aeruginosa to respond to circulating neutrophils that produce RCS at much lower concentrations. At the same time, flow could amplify the bactericidal effects against P. aeruginosa. Future work will need to determine the relative steady-state abundances of different secondary RCS and address their impact in combination with flow on bactericidal activity and the activation of the Fro response.

At the fundamental level, Fro is a system that increases antimicrobial tolerance, which can promote antimicrobial resistance. For example, bacteria that are exposed to low concentrations of HOCl have higher rates of horizontal gene transfer and can acquire antibiotic resistance genes more rapidly (Hou et al., 2019; Jin et al., 2020). The Fro system could thus facilitate antibiotic resistance through increasing P. aeruginosa tolerance to HOCl. Our findings suggest that the bacterial HOCl response could thus be an important bacterial defense mechanism to target in the development of novel antibiotic therapeutics. Inhibition of this pathway could increase P. aeruginosa susceptibility to neutrophil-mediated killing while minimizing antibiotic resistance.

Materials and methods

Bacterial Culture Conditions

Strains were struck out from frozen stocks maintained at -80°C onto plates containing LB-Miller broth and 2% Bacto agar and incubated overnight at 37°C. Single colonies were inoculated into 2mL of modified MinA minimal medium (60.3mM K2HPO4, 33.0mM KH2PO4, 7.6mM (NH4)2SO4, 1.0mM MgSO4; (Malke, 1993) containing 0.2% (w/v) sodium citrate (hereafter referred to as MinA minimal medium), and cultured for 18 hours in a roller drum at 17 rpm and 37 °C.

Imaging P. aeruginosa under flow

Microfluidic devices were fabricated using standard soft lithography techniques as established previously (Siryaporn et al., 2015) with channel dimensions of 100 x 50 μm (width x height). P. aeruginosa was cultured for 18 hours in LB-miller medium, diluted 1:100, cultured to mid-exponential phase, and injected into the microfluidic device. Fresh LB was injected at a constant rate of 10μL/min. Images were acquired as described in the fluorescence microscopy section.

Neutrophil Isolation

Neutrophils were isolated freshly from whole blood as described in (Clark et al., 2018). Briefly, 3% dextran in PBS was used to separate red blood cells (RBCs) from whole blood. Neutrophils were purified from the remaining cells by overlaying on a Ficoll density gradient and centrifugation in for 30 minutes at 500 × g. RBCs were lysed using RBC Lysis Buffer and centrifugation for 5 mins at 300 × g. Supernatant was aspirated and the remaining neutrophil pellets were resuspended in RPMI 1640 medium prior to normalizing cell counts. Neutrophils were used immediately for stimulation, preparation of conditioned media, or co-incubation experiments with P. aeruginosa as described below, and purity was assessed in parallel using the fluorescent antibodies APC-CD11b, FITC-CD16 and PE-CD66b and a ACEA NovoCyte Flow Cytometer. This procedure routinely yielded greater than 95% CD11b-positive neutrophil populations.

Co-incubation of Mammalian Cells with P. aeruginosa

Neutrophils were prepared as described above. J774A.1 mouse macrophages were cultured in 10 mL DMEM in T75 flasks at 37 °C with 5% CO2. Macrophages were passaged every three days through scraping and were passaged every 5 days through trypsinization. Mammalian cells were washed three times by centrifugation at 300 × g for 5 minutes and resuspension in DPBS, and 1 mL aliquots of 4x105 cells were transferred into flat-bottom dishes. P. aeruginosa overnight (∼18h) cultures were centrifuged at 4600 × g for 2 minutes, resuspended in DPBS three times, diluted to a final OD600 of 0.2, added to the mammalian cells at a multiplicity of infection of 10, and incubated for 30 minutes at 37°C. Dishes were aspirated until approximately 20 µL of medium remained. 1% agarose pads containing DPBS were placed on top of the samples in the dishes to immobilize bacteria and cells. Samples were imaged as described in the fluorescence microscopy section.

Conditioned Medium from Neutrophils

To activate neutrophils, 2 million freshly harvested neutrophils per mL in 10mM phosphate buffer (pH 7.4; 2.46 mM monobasic phosphate, 7.54 mM dibasic sodium phosphate) containing 140 mM sodium chloride, 10 mM potassium chloride, 0.5 mM magnesium chloride, 1 mM calcium chloride, 1 mg/mL glucose, and 5 mM taurine (Dypbukt et al., 2005), were treated with 100 ng/mL of PMA in 14mL polystyrene round-bottom Falcon tubes for 1 hour at 37°C and 17 rpm. For MPO inhibition studies, neutrophils were pre-treated with 2 mM of 4-ABAH for 30 mins prior to addition of PMA. This concentration of 4-ABAH was scaled up based on its usage of 500 μM of 4-ABAH to treat one million human neutrophils per mL (Tseng et al., 2018). Supernatant was collected after centrifugation for 5 minutes at 4,600 × g and 24 °C and used as conditioned medium immediately for HOCl detection assay as described below. In parallel, P. aeruginosa that were cultured for 18 hours in MinA minimal medium were diluted 1:1000 into conditioned media and were incubated for 3 hours at 37 °C in a rolling drum at 17 rpm and imaged as described in the fluorescence microscopy section.

HOCl Fluorometric Detection Assay

Conditioned medium was assessed immediately using the Amplite Fluorometric Hypochlorite Assay Kit according to manufacturer protocol. Samples were incubated at room temperature in the dark for 10 minutes and measured for fluorescence using a BioTek Synergy HTX plate with 540/25 nm and 585/10 nm filters for excitation and emission, respectively, and using the autogain set to the 1 μM NaOCl condition. HOCl concentrations were determined by fitting the calibration values to the Hill equation using a three-parameter fit and the lowest value in the data series as the minimum value. HOCl concentrations were background-subtracted by the neutrophil buffer-only sample.

Treatments of P. aeruginosa with oxidizers, antioxidants, and stressors

P. aeruginosa were cultured 18 hours in MinA minimal medium and diluted 1:1000 into the same medium. HOCl, H2O2, HNO3, NaOH, NaCl, methionine, L-cysteine, or β-mercaptoethanol (βME) were added to 50 mL cultures at the concentrations indicated in the main text and were incubated in 250 mL flasks for 3 hours at 37 °C in an orbital shaker at 225 rpm. LL-37, histones and tobramycin were added following the 1:1000 diluted and incubated in 14 mL falcon tubes in a roller drum at 17 rpm. Cultures were imaged and analyzed as described in the fluorescence microscopy section.

Fluorescence and Phase Contrast Microscopy

P. aeruginosa were immobilized on 1% agarose pads containing MinA minimal medium and imaged immediately. Cultures containing densities below 10 P. aeruginosa per frame were concentrated using a syringe filter and imaged immediately. Experiments in Figures 2D and 3F were performed without the syringe filter in order to confirm that the P. aeruginosa response was not affected by the filtering procedure.

Imaging

Images were acquired using Nikon NIS-Elements software and a Nikon Eclipse Ti-E microscope containing a Nikon 100X Plan Apo (1.45 N.A.) objective, Nikon Ph3 phase contrast condenser annulus, Sola light engine, Hamamatsu Orca Flash 4.0 V2 camera, 459/526/596 dichroic, 575/25 excitation and 641/75 emission filters for mCherry, and 509/22 excitation and 544/24 emission filters for YFP (Semrock, Rochester, NY).

Image Analysis

Images were analyzed in MATLAB using custom built software written previously (see the Key Resources Table to download). Briefly, P. aeruginosa masks were determined from phase contrast images using an edge-detection algorithm. For size analysis, the total pixel area of each individual P. aeruginosa cell was determined by computing the mask area and converting from pixels to μm2 by multiplying the mask area by a factor of 0.004225 μm2/pixel to account for the microscope camera pixel size and objective magnification. Fro expression was quantified by averaging computed ratios of YFP to mCherry fluorescence from the masked areas for at least one hundred individual P. aeruginosa cells per condition for each experiment.

Bacterial Growth Curves

Overnight cultures of bacteria were diluted into MinA minimal medium to an optical density at 600 nm (OD600) of 0.01, given treatments, and incubated in sterile 96-well microplates at 37°C in a Synergy HTX multi-mode plate reader with continuous orbital shaking at 180 rpm and a 3mm amplitude, with OD600 readings taken every 30 minutes.

Reverse Transcription Quantitative PCR (RT-qPCR)

For non-murine model samples, bacterial overnight cultures were diluted 1:1000 into fresh MinA medium and incubated for 2.5 hours at 37°C and 225 rpm preceding a 30 minute treatment with 1 µM NaOCl or vehicle control. Samples were concentrated using syringe filters and centrifuged at 13,000 × g for 4 minutes. Pellets were snap-frozen in liquid nitrogen and stored at -80C before performing RNA extraction with the NucleoSpin RNA kit. cDNA was prepared from the using the High-Capacity cDNA Reverse Transcriptase Kit, and quantitative PCR was performed using the SsoAdvanced Universal SYBR Green Supermix using the C1000 Thermal Cycler with a CFX96 real-time detection system. froA transcripts were quantified using the primers PA14_froA_v2_F and PA14_froA_v2_R that target PA14_21570 in PA14 and PA3284 in PAO1 (see Key Resources Table). Transcript abundances for each sample were normalized by 5S rRNA abundance using the primers PAO1_PA14_5S_F and PAO1_PA14_5S_R (see Key Resource Table), yielding normalized count thresholds (Cts). The average fold change in transcript abundance was determined by computing the ratio of the averaged normalized Cts and raising by the power of 2, as performed previously (Bru et al., 2019). All primers were validated using a standard calibrating curve to assess Cts and transcript abundance.

Murine Model of Corneal Infection

Expression of froA was assessed in a murine model of corneal infection as described previously (Minns et al., 2023; Sun et al., 2012). The work received prior approval from the Institutional Animal Care and Use Committee at the University of California Irvine under protocol #AUP-21-123. Briefly, P. aeruginosa strain PAO1F was cultured to mid-exponential phase in MinA minimal medium that was modified to contain only 0.05% citrate and supplemented with 0.1% casamino acids and 0.2% glucose as the carbon sources, centrifuged, and resuspended in PBS. Corneal epithelia of C57BL/6J mice aged 6-8 weeks were abraded with 3x5 mm scratches using a 25G needle, and 2 µL of PBS containing 5x104 of P. aeruginosa strain PAO1F was applied topically. Mice were selected at random and those displaying signs of illness or distress were excluded from the study. Comparable numbers of male and female mice were used in each experimental group. The number of animals used for each experiment was determined based on a power analysis estimate. Mouse eyeballs were collected and homogenized in 1 mL of PBS 2 or 20 hours post-infection and centrifuged for 5 minutes at 150 × g to remove corneal materials. Supernatant was collected and centrifuged for 4 minutes at 13,000 × g. Pellets containing bacteria were suspended in lysis solution (10 mM Tris-HCl, 1 mM EDTA pH 8.0, 0.5 mg/mL lysozyme, 1% SDS) (Bru et al., 2019) RNA was prepared using the NucleoSpin RNA kit and assessed for mRNA expression using RT-qPCR. Pellets containing bacteria were suspended in lysis solution (10 mM Tris-HCl, 1 mM EDTA pH 8.0, 0.5 mg/mL lysozyme, 1% SDS) (Bru et al., 2019).

Samples were not blinded because they were processed as they were collected at different timepoints, which revealed their identity.

RNA-seq Library Preparation and Data Analysis

P. aeruginosa was cultured and RNA was prepared as described in the RT-qPCR section. RNA yield was measured using a Nanodrop 2000. Samples were depleted of ribosomal RNA using the NEBNext rRNA Depletion kit, from which cDNA libraries were constructed using the NEBNext Ultra II Directional Library kit, which were sequenced by the UC Irvine Genomics Research & Technology Hub (Irvine, CA) using an Illumina NovaSeq X Plus using paired-end 150 bp reads at a depth of approximately 10 million reads per sample. Raw reads were checked with fastQC and trimmed and filtered into paired and unpaired reads using Trimmomatic in Java using the ‘PE’ setting. Paired and unpaired reads were aligned to the PA14 genome (NCBI accession NC_008463.1) using Bowtie2 using the default ‘sensitive’ mode, were counted using the featureCounts program in Subread with the ‘countReadPairs’ option enabled for paired-end reads, and were fit into a negative binomial model to compute fold-change in gene expression using DeSeq2. The statistical significance of the changes were computed using the Wald test in DeSeq2. Gene Ontology enrichment analysis for Figure 1A was performed using GOEnrichment.

Statistical Analyses

Statistical analyses and figures were generated using Excel, R, and RStudio.